---
title: "Assignment 8"
output: github_document

---


# Homework Assignment: Handwritten Letter Recognition with Neural Networks

## Overview

In this assignment you will build, evaluate, and regularize a neural network to classify handwritten letters using the EMNIST dataset.
Where lectures 14–16 used the first ten digits (0–9), here you will work with the **first ten letters of the alphabet (a–j)** from the EMNIST letters dataset.
The pixel structure is identical — each observation is a 28×28 grayscale image flattened to 784 features — but the classification problem is different: letters share more visual features than digits (think of b, d, p, and q), making this a meaningfully harder problem.

The assignment covers the full arc of the neural network unit: data structure and visualization, backpropagation intuition, model fitting in tidymodels, and regularization.

---

## Step 1: Data Preparation and Visualization

### 1.1 Load the data

Load the EMNIST letters training file (`emnist-letters-train.csv`).
Filter to retain only the first ten letter classes (labels 1–10, corresponding to letters a–j).
Rename the columns so the first column is called `label` and the remaining 784 columns are named `pix_1` through `pix_784`, consistent with the convention used in lecture.

### 1.2 Recode the label

The EMNIST letters file uses numeric codes 1–26 for a–z.
Create a new factor variable `letter` that maps codes 1–10 to the corresponding lowercase letters ("a" through "j").
This will serve as your outcome variable throughout the assignment.

### 1.3 Visualize sample letters

Adapt the `plot_emnist_digit()` function from lecture 14 to work with this dataset.
Call your version `plot_emnist_letter()`.
Use it to display at least **two examples of each of the ten letters** (20 images minimum).

For at least one visually similar pair (e.g., b and d, or i and j), display the two letters side by side and comment in 2–3 sentences on what visual features a neural network might need to distinguish them.


## Step 3: Fit a Neural Network in tidymodels

Fit a neural network to the  EMNIST letters data using the tidymodels / brulee workflow from lecture 16.

### 3.1 Split the data

Create a stratified 80/20 train/test split on the `letter` variable.
Create 5-fold cross-validation folds from the training set, also stratified on `letter`.


### 3.2 Build the recipe

Build a tidymodels recipe that:

- Sets `letter` as the outcome and all `pix_*` columns as predictors
- Normalizes all predictors to mean 0 and standard deviation 1


### 3.3 Model specification — dropout workflow

Following the lecture 16 approach, fit a dropout-regularized model.
Use `mlp()` with `set_engine("brulee")`.
Fix `penalty = 0` and tune `hidden_units`, `epochs`, and `dropout`.

Build a Latin hypercube tuning grid with at least 10 parameter combinations.
Tune using your CV folds with ROC AUC (macro-weighted) and accuracy as metrics.

### 3.4 Model specification — L2 penalty workflow

Following the same structure, fit a second model tuning `penalty` with `dropout = 0` fixed.
Tune `hidden_units`, `epochs`, and `penalty`.
Use the same grid size and CV folds.

### 3.5 Inspect tuning results

For each tuning workflow, produce:

1. A ranked plot of cross-validated ROC AUC across all grid points (with standard error bars)
2. The output of `show_best()` for ROC AUC
3. The selected best parameter configuration from `select_best()`

## Step 4: Final Model and Evaluation

### 4.1 Select and finalize

Select the better-performing regularization strategy (dropout or L2) based on cross-validated ROC AUC.
Finalize the workflow with the best parameters and use `last_fit()` to fit on the full training set and evaluate on the test set.

### 4.2 Performance metrics

Report overall accuracy, ROC AUC (macro-weighted), and mean log loss on the test set.
Compare accuracy to the null classifier baseline (always predicting the most common letter in the training set).

```{r metrics}
# Your code here — collect_metrics(final_res)
# Compute null classifier baseline from training data
```

**Discussion (2–3 sentences):** How does the model perform relative to the null classifier?
Is this level of accuracy sufficient for a real-world handwriting recognition application?
Why or why not?

### 4.3 Confusion matrix

Produce a confusion matrix heatmap for the test set predictions.
Label the axes with the actual letter names (a–j), not numeric codes.

```{r confusion-matrix}
# Your code here
```

**Discussion:** Identify the two or three most commonly confused letter pairs in the confusion matrix.
In 3–4 sentences, explain why those specific pairs are confused and what this tells you about the features the network is (or isn't) learning.

---

## Step 5: Regularization Effects

### 5.1 Simulate regularization on synthetic data

Using the `train_network()` function from lecture 16, train three versions of a network on the synthetic 10-class, 784-input data from Step 2:

1. No regularization (`lambda = 0`, `drop_rate = 0`)
2. L2 regularization (`lambda = 0.01`, `drop_rate = 0`)
3. Dropout regularization (`lambda = 0`, `drop_rate = 0.4`)

```{r simulate-reg}
# Your code here — copy train_network() from lecture 16
```

### 5.2 Weight distribution plot

Reproduce the W1 weight histogram from lecture 16, comparing the three regularization conditions.

```{r weight-hist}
# Your code here
```

**Discussion (2–3 sentences):** How do L2 regularization and dropout affect the distribution of learned weights differently?
Which condition produces the narrowest weight distribution, and why does L2 have this effect mathematically?

### 5.3 Weight magnitude by layer

Reproduce the mean absolute weight magnitude bar chart from lecture 16, comparing all three conditions across both layers (W1 and W2).

```{r weight-magnitude}
# Your code here
```

**Discussion (2–3 sentences):** Does regularization affect W1 and W2 similarly, or does it have a stronger effect on one layer?
What does this suggest about where in the network regularization has the most impact?

---

## Step 6: Conceptual Questions

Answer each question in 3–5 sentences.

**6.1** Explain in your own words why dropout acts as an implicit ensemble method.
What is the connection between randomly silencing neurons during training and averaging over many different network architectures?

**6.2** In lecture 15 we saw that gradients in layers closer to the input tend to be smaller than gradients in layers closer to the output.
Why does this happen, and what is the term for the extreme version of this phenomenon in very deep networks?

**6.3** The EMNIST letters task (a–j) is generally considered harder than the digits task (0–9).
Based on what you observed in the confusion matrix and the visualizations in Step 1, give a specific mechanistic reason — in terms of pixel-level features — for why letter classification is more difficult.

**6.4** Cross-entropy loss heavily penalizes confident but incorrect predictions.
A model that assigns 1% probability to the correct class incurs a loss of approximately 4.6, while a model that assigns 95% probability to the correct class incurs a loss of approximately 0.05.
Why is this asymmetry desirable when training a classifier?
What would go wrong if we used mean squared error instead?

---

## Submission Checklist

Before submitting, confirm that your rendered document includes:

- [ ] Pixel visualization for all 10 letters (at least 2 examples each)
- [ ] All four backpropagation visualizations from Step 2 with written interpretations
- [ ] Tuning results for both the dropout and L2 workflows
- [ ] Confusion matrix with letter labels on both axes
- [ ] Weight distribution histogram and magnitude bar chart
- [ ] Written responses to all four conceptual questions
- [ ] `fit_model <- FALSE` set for all expensive chunks so the document renders quickly
