---
title: "Introduction to Tidyverse/Tidymodels"
author: "Will Doyle"
date: "2025-01-07"
output: github_document
---

## Datset

The dataset for this session comes from HSLS, which is nationally representative sample of ninth graders as of 2009. The data combines questionnaire information and transcript information.

## Outcome of interest

We want to predict total HS GPA from a set of characteristics of incoming ninth graders. The use case here is quickly identifying those students that may end up with low GPAs so they can be targeted for early interventions. 

## Measure of accuracy

We'll use root mean sqaure error (rmse) as our measure of accuracy. 

## Load libraries
```{r}
library(tidyverse)
library(tidymodels)
library(janitor)
```


## Load dataset
```{r}
hs<-read_csv("hsls_extract.csv")%>%clean_names()
```

## Codebook

Categorical/factor variables marked with (f)

x3gpatot: Final HS GPA
x1region: Region of the country (f)
x1locale: Locale (f)
x1control: Control (f)
x1iepflag: IEP Flag (f)
x1stuedexpct: Student expectations for educational attainment (f)
x1schooleng: Student engagement
x1famincome: Family income (f)
x1hhnumber: Number in household (f)
x1par1emp: Parent's employment status (f)
x1par1edu: Parent's education level (f)
x1tmtscor: Math theta score in ninth grade

## Data Cleaning

I need to properly code missing data, and then drop any cases with missing data. 
This won't always be our approach.

```{r}
hs <- hs %>%
  mutate(across(-x1txmtscor, ~ ifelse(. < 0, NA, .)))%>%
  drop_na()
```


`tidymodels` is a collection of R packages that provides a consistent and tidy approach to building, evaluating, and tuning statistical models. It integrates various stages of the modeling process, from data preprocessing to model evaluation, under a unified framework that's easy to use and aligns with the principles of the `tidyverse`.

 **Philosophy**:
 
   - **Tidy Data Principles**: `tidymodels` is built on the same principles as the `tidyverse`, emphasizing clear, readable, and consistent code. It's designed to work seamlessly with other `tidyverse` packages.
   
   - **Unified Interface**: It offers a consistent interface for various modeling tasks, regardless of the underlying model type.

 **Components**:
 
   - **`recipes`**: Provides tools for feature engineering and data preprocessing. It allows you to specify a series of steps to prepare your data for modeling.
   
   - **`parsnip`**: A unified interface for model specification. It allows you to define a model without committing to a specific computational engine.
   
   - **`workflows`**: Combines pre-processing steps from `recipes` and model specifications from `parsnip` into a single object to streamline the modeling process.
   
   - **`tune`**: Tools for model tuning, such as grid search and cross-validation.
   
   - **`yardstick`**: For model evaluation and metric calculation.
   
   - **`dials`**: Helps in creating tuning grids for different model parameters.
   
   - **`rsample`**: Provides infrastructure for data splitting and resampling, which is essential for model validation and testing.

**Workflow**:

   - **Data Splitting**: Using `rsample`, you can create training and testing datasets.
   
   - **Preprocessing**: With `recipes`, you define a series of preprocessing steps like normalization, encoding categorical variables, and more.
   
   - **Model Specification**: Using `parsnip`, you specify the type of model you want (e.g., linear regression, decision tree) without choosing the computational engine (e.g., lm, xgboost).
   
   - **Combining Preprocessing and Modeling**: `workflows` lets you combine your preprocessing steps and model specification into a single object.
   
   - **Model Training**: Train your model on the training data.
   
   - **Model Evaluation**: Once trained, you can evaluate your model's performance on the testing data using `yardstick`.
   
   - **Tuning (if needed)**: If your model has hyperparameters, you can use `tune` and `dials` to find the best parameters.

 **Advantages**:
 
   - **Flexibility**: Easily switch between different model types or computational engines without drastically changing your code.
   
   - **Consistency**: Regardless of the model type, the code structure remains consistent.
   
   - **Integration**: Designed to work seamlessly with other `tidyverse` packages, making it easier to integrate modeling into a broader data analysis pipeline.

Below I'll use tidymodels to run a regresion predicting overall HS GPA from the incoming student characteristics. 

## Splitting into training and testing

The code is dividing the `cr` dataset into two parts: a training set (`train`) to teach a machine learning model and a testing set (`test`) to evaluate how well the model has learned. This split ensures that the model is evaluated on data it hasn't seen before, providing a more realistic assessment of its performance.

```{r}
hs_split<-initial_split(hs)

train<-training(hs_split)

test<-testing(hs_split)
```

 **`hs_split <- initial_split(hs)`**:
 
   - This line uses the `initial_split` function to divide the `cr` dataset into two parts. By default, `initial_split` typically allocates 75% of the data to the training set and the remaining 25% to the testing set, though this ratio can be adjusted.
   
   - The result, `cr_split`, is a special split object that contains information about which rows of the original `cr` dataset belong to the training set and which belong to the testing set.

 **`train <- training(hs_split)`**:
 
   - Here, the `training` function extracts the training portion of the data from the `cr_split` object and assigns it to the `train` variable. This dataset will be used to train a machine learning model.

 **`test <- testing(hs_split)`**:
 
   - Similarly, the `testing` function extracts the testing portion of the data from the `cr_split` object and assigns it to the `test` variable. This dataset will be used to evaluate the performance of the trained model on unseen data.


The code is setting up a linear regression model using the standard linear modeling engine in R.
```{r}
hs_model<-linear_reg(mode="regression",engine="lm")
```


**Detailed Explanation**:

- **`linear_reg(mode="regression", engine="lm")`**:

  - **`linear_reg()`**: This function is from the `parsnip` package, which is part of the `tidymodels` framework. It's used to specify a linear regression model.
  
  - **`mode="regression"`**: This argument specifies the type of modeling. In this case, it's regression, which is used to predict a continuous outcome variable based on one or more predictor variables.
  
  - **`engine="lm"`**: This argument specifies the computational engine to use for the linear regression. The "lm" engine refers to R's built-in `lm()` function for linear modeling.

- **`hs_model`**: The specified linear regression model is then stored in the `hs_model` variable. This doesn't train the model yet; it merely sets up the type of model and the engine to be used. Training the model would require additional steps using the training data.

## Set Recipe

In the tidymodels framework, recipes is a package that provides a streamlined way to define and preprocess data for modeling. It allows users to specify a series of steps to transform and preprocess data, such as normalization, encoding categorical variables, and handling missing values. These steps are defined in a consistent and reproducible manner, creating a "recipe" that can be applied to both training and new datasets. This ensures that data transformations are consistent across different stages of the modeling process, facilitating more reliable model training and prediction. Essentially, recipes offers a tidy and systematic approach to data preparation in the modeling workflow.

The code below is preparing the data for modeling. It sets up a series of steps to process the data, such as handling missing values, converting categorical variables into a format suitable for modeling, and normalizing the data.

```{r}
hs_formula<-as.formula("x3tgpatot~.")

hs_rec<-recipe(hs_formula,data=train)%>%
  update_role(x3tgpatot,new_role = "outcome")%>%
  step_other(all_nominal_predictors(),threshold = .01)%>%
  step_dummy(all_nominal_predictors())%>%
  step_filter_missing(all_predictors(),threshold = .1)%>%
  step_naomit(all_outcomes(),all_predictors())%>%
  step_corr(all_predictors(),threshold = .95)%>%
  step_zv(all_predictors())%>%
  step_normalize(all_predictors())
```

The code is preparing the data for modeling. It sets up a series of steps to process the data, such as handling missing values, converting categorical variables into a format suitable for modeling, and normalizing the data.

**Detailed Explanation**:

1. **`hs_formula <- as.formula("x3gpatot~.")`**:

   - This creates a formula indicating that the variable `x3gpatot` is the outcome (or dependent variable) we want to predict, and the `.` means we want to use all other variables in the dataset as predictors (or independent variables).

2. **`recipe(hs_formula, data=train)`**:

   - The `recipe` function starts the specification of preprocessing steps. It uses the formula and the training data (`train`) as inputs.

3. **`update_role(x3gpatot, new_role = "outcome")`**:

   - This explicitly sets the role of the `x3gpatot` variable as the "outcome" or the variable we're trying to predict.

4. **`step_other(all_nominal_predictors(), threshold = .01)`**:

   - For categorical (nominal) predictors, any categories that constitute less than 1% of the data will be lumped together into a new category, typically called "other".

5. **`step_dummy(all_nominal_predictors())`**:

   - Converts categorical variables into dummy variables (also known as one-hot encoding). This is necessary because many modeling algorithms require numerical input.

6. **`step_filter_missing(all_predictors(), threshold = .1)`**:

   - This step removes any predictor variables that have more than 10% missing values.

7. **`step_naomit(all_outcomes(), all_predictors())`**:

   - Removes rows (observations) from the data where either the outcome or any of the predictor variables have missing values.

8. **`step_corr(all_predictors(), threshold = .95)`**:

   - Identifies and removes predictor variables that have a correlation higher than 0.95 with any other predictor. This helps in addressing multicollinearity.

9. **`step_zv(all_predictors())`**:

   - Removes predictor variables that have a zero variance, meaning they have the same value for all observations.

10. **`step_normalize(all_predictors())`**:

   - Normalizes all predictor variables so they have a mean of 0 and a standard deviation of 1. We'll do this for everything, it's a fairly standard step. 

The result of all these steps is stored in `hs_rec`, which can then be used to preprocess the training data and any future data in a consistent manner before modeling.


## Viewing the results of feature engineering
```{r}
hs_rec%>%prep()

hs_rec%>%prep()%>%bake(train)
```

1. **`hs_rec %>% prep()`**:

   - **`prep()`**: This function prepares the steps defined in the `cr_rec` recipe. It computes any required statistics or parameters needed for the transformations (e.g., mean and standard deviation for normalization) but doesn't apply the transformations to the data yet. Think of it as getting the recipe ready for cooking but not actually cooking.

2. **`hs_rec %>% prep() %>% bake(train)`**:

   - **`bake()`**: Once the recipe is prepared with `prep()`, the `bake()` function is used to apply the transformations to a dataset. In this case, the transformations are applied to the `train` dataset.
   
   - Essentially, this sequence of functions means "prepare the transformations and then apply them to the `train` dataset."

By separating the `prep()` and `bake()` steps, `tidymodels` allows for a consistent set of transformations to be easily applied to different datasets, ensuring that both training and testing data undergo the same preprocessing.


## Creating a workflow and fitting the model
```{r}
hs_wf<-workflow()%>%
  add_model(hs_model)%>%
  add_recipe(hs_rec)%>%
  fit(train)
```

The code is setting up a workflow that combines both the data preprocessing steps (from `hs_rec`) and the modeling specification (from `hs_model`). It then trains the model using the `train` dataset.


1. **`workflow()`**:

   - This function initializes a workflow. In the `tidymodels` framework, a workflow is a way to bundle together preprocessing steps (like those defined in a recipe) and a model specification.

2. **`add_model(cr_model)`**:

   - This adds the model specification from `hs_model` (which was defined earlier) to the workflow. It tells the workflow what kind of model we're planning to train.

3. **`add_recipe(cr_rec)`**:

   - This adds the data preprocessing steps from `hs_rec` (which were defined earlier) to the workflow. It tells the workflow how to preprocess the data before training the model.

4. **`fit(train)`**:

   - Once the workflow has both the model and the recipe, the `fit()` function is used to train the model using the `train` dataset. This involves applying the preprocessing steps to the training data and then training the model on the processed data.

The result is stored in `hs_wf`. This object now contains the trained model along with all the preprocessing steps, making it ready for predictions on new data.

## Testing predictions in the testing split

```{r}
test<-
  hs_wf%>%
  predict(new_data=test)%>%
  bind_cols(test)
```


```{r}
test%>%
  rmse(truth="x3tgpatot",estimate=.pred)
```

The code calculates the Root Mean Squared Error (RMSE) for the predictions in the `test` dataset by comparing the predicted values (`.pred`) with the actual values (`x3tgpatot`).

**Detailed Explanation**:

1. **`test %>%`**:

   - This takes the `test` dataset, which now contains both the actual values (`x3tgpatot`) and the predicted values (`.pred`), as the starting point for the subsequent operations.

2. **`rmse(truth="iss", estimate=.pred)`**:

   - The `rmse()` function from the `yardstick` package (part of `tidymodels`) is used to calculate the RMSE.
   
   - **`truth="x3tgpatot"`**: Specifies that the actual values are in the `x3tgpatot` column of the `test` dataset.
   
   - **`estimate=.pred`**: Specifies that the predicted values are in the `.pred` column of the `test` dataset.

The result will be the RMSE value, which provides a measure of the differences between the predicted and actual values. A lower RMSE indicates a better fit of the model to the data, while a higher RMSE suggests a poorer fit.

Note: Ensure that the `yardstick` package is loaded to use the `rmse()` function.


## Examining Coefficients

Coefficients? Who cares about coefficients? We've already got an rmse!

The code fits the model above to the full dataset, extracts the coefficients from the trained model from the workflow, organizes these results in a tidy format, sorts them based on the magnitude of the estimates (from highest to lowest), and then displays the top 100 results.

```{r}
hs_wf_full<-
  hs_wf%>%
  fit(hs)

hs_wf_full%>%
  extract_fit_parsnip()%>%
  tidy()%>%
  arrange(-abs(estimate))%>%
  print(n=100)
```


**Detailed Explanation**:

1. **`hs_wf %>%`**:

   - This takes the trained workflow `cr_wf` as the starting point for the subsequent operations.

2. **`extract_fit_parsnip()`**:

   - This function extracts the results of the trained model from the workflow. In the context of a linear regression model (as indicated by previous interactions), this would typically include coefficients for each predictor variable.

3. **`tidy()`**:

   - This function, from the `broom` package, tidies the results of the model, converting them into a clean and standardized data frame format. For a linear regression model, this would typically result in a data frame with columns like `term` (the predictor variable name), `estimate` (the coefficient value), and others like `std.error`, `statistic`, and `p.value`.

4. **`arrange(-estimate)`**:

   - This arranges (or sorts) the results based on the `estimate` column in descending order (from highest to lowest coefficient value).

5. **`print(n=100)`**:

   - This displays the top 100 results after sorting. It's especially useful if there are many predictor variables in the model and you want to see the ones with the highest coefficient values.


## Moving forward

We'll next use lasso, ridge and elastic net to select the features that contribute the most to the model, resulting in a model that should be able to better predict the outcome in the testing data. 




