---
title: "Backpropagation"
author: "Will Doyle"
date: "2025-03-25"
output: html_document
---

## Backpropagation

In this lesson, we'll work through a simulated example to show how backpropagation works. This is somewhat abstracted from the actual way neural nets function, but close enough to give you a good picture of what's going on.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
```

---

## Synthetic Data

The function below, `generate_synthetic_data`, creates a synthetic dataset for a multi-class classification problem. It allows the user to specify the number of samples (`n_samples`), the dimensionality of each input vector (`input_dim`), the number of classes (`n_classes`), and the standard deviation of noise (`noise_sd`). It generates random centroids — one per class — then for each sample selects a class and adds Gaussian noise around that class's centroid. Inputs are stored in matrix `X` and labels are one-hot encoded in matrix `Y`.

```{r generate-data}
generate_synthetic_data <- function(n_samples = 1000,
                                    input_dim = 10,
                                    n_classes = 5,
                                    noise_sd = 0.1) {
  centroids <- matrix(
    rnorm(n_classes * input_dim, mean = 0, sd = 1),
    nrow = n_classes, ncol = input_dim
  )

  class_ids <- sample(0:(n_classes - 1), n_samples, replace = TRUE)

  X <- centroids[class_ids + 1, ] +
    matrix(rnorm(n_samples * input_dim, 0, noise_sd), n_samples, input_dim)

  Y <- matrix(0, nrow = n_samples, ncol = n_classes)
  Y[cbind(seq_len(n_samples), class_ids + 1)] <- 1

  list(X = X, Y = Y)
}
```

```{r set-dims}
k <- 784   # input dimension
n <- 1000  # samples
j <- 10    # classes

set.seed(42)
synth <- generate_synthetic_data(
  n_samples  = n,
  input_dim  = k,
  n_classes  = j,
  noise_sd   = 0.5
)

X <- synth$X
Y <- synth$Y
```

---

## Visualization 1: PCA of Input Space

Before training anything, it helps to see whether the synthetic classes are separable in the input space. We project onto principal components 3 and 4 to illustrate that classes form loose clusters.

```{r pca-plot}
labels <- apply(Y, 1, function(row) which(row == 1) - 1)

prcomp(X, center = TRUE, scale. = TRUE)$x[, 3:4] %>%
  as_tibble(.name_repair = "minimal") %>%
  set_names(c("PC3", "PC4")) %>%
  mutate(class = as.factor(labels)) %>%
  ggplot(aes(x = PC3, y = PC4, color = class)) +
  geom_point(alpha = 0.6, size = 1.5) +
  labs(
    title = "PCA projection of synthetic input data",
    x = "Principal Component 3",
    y = "Principal Component 4",
    color = "Class"
  ) +
  theme_minimal()
```

---

## Network Initialization

The `initialize_network` function sets up a two-layer neural network by randomly initializing weights and biases. Weights are drawn uniformly from [-0.1, 0.1]; biases start at zero.

```{r initialize}
initialize_network <- function(input_dim = k, hidden_dim = 5, output_dim = j) {
  list(
    W1 = matrix(runif(hidden_dim * input_dim, -0.1, 0.1), nrow = hidden_dim),
    b1 = rep(0, hidden_dim),
    W2 = matrix(runif(output_dim * hidden_dim, -0.1, 0.1), nrow = output_dim),
    b2 = rep(0, output_dim)
  )
}

network_init <- initialize_network()
```

---

## Forward and Backward Pass

The `backprop_step` function performs one full forward pass, computes cross-entropy loss, runs backpropagation to compute gradients, and returns updated weights.

```{r backprop-function}
softmax <- function(z) {
  exp_z <- exp(z - apply(z, 1, max))
  exp_z / rowSums(exp_z)
}

backprop_step <- function(X, Y, network, lr = 0.1) {
  n_samples <- nrow(X)

  W1 <- network$W1
  b1 <- network$b1
  W2 <- network$W2
  b2 <- network$b2

  # --- FORWARD PASS ---
  Z1 <- X %*% t(W1) + matrix(rep(b1, each = n_samples), nrow = n_samples)
  A1 <- pmax(0, Z1)
  A1 <- matrix(A1, n_samples, nrow(W1))

  Z2    <- A1 %*% t(W2) + matrix(rep(b2, each = n_samples), nrow = n_samples)
  Y_hat <- softmax(Z2)
  loss  <- -sum(Y * log(Y_hat + 1e-12))

  # --- BACKWARD PASS ---
  dZ2 <- (Y_hat - Y) / n_samples
  dW2 <- t(dZ2) %*% A1
  db2 <- colSums(dZ2)

  dA1       <- dZ2 %*% W2
  dZ1       <- dA1
  dZ1[Z1 <= 0] <- 0

  dW1 <- t(dZ1) %*% X
  db1 <- colSums(dZ1)

  list(
    network = list(
      W1 = W1 - lr * dW1,
      b1 = b1 - lr * db1,
      W2 = W2 - lr * dW2,
      b2 = b2 - lr * db2
    ),
    prediction = Y_hat,
    loss       = loss,
    dW1_norm   = norm(dW1, "F"),
    dW2_norm   = norm(dW2, "F")
  )
}
```

---

## Training Loop

We run 1,000 iterations using `accumulate()` from purrr, which threads the network state forward at each step and retains the full history.

```{r train}
training_history <- accumulate(
  1:1000,
  function(state, i) backprop_step(X, Y, state$network, lr = 0.5),
  .init = list(network = network_init, loss = NA, dW1_norm = NA, dW2_norm = NA)
)

# Extract scalar history into a tidy data frame (drop .init step at index 1)
history_df <- training_history[-1] %>%
  map_dfr(~ tibble(loss = .x$loss,
                   dW1_norm = .x$dW1_norm,
                   dW2_norm = .x$dW2_norm)) %>%
  mutate(iteration = row_number())

# Trained network is the final state
network_trained <- training_history[[1001]]$network
```

---

## Visualization 2: Loss Curve

The loss should decrease monotonically (given a suitable learning rate). The dashed line marks the iteration with the steepest single-step drop — the moment of fastest learning.

```{r viz-loss}
steepest <- history_df %>%
  mutate(loss_drop = loss - lag(loss)) %>%
  slice_min(loss_drop, n = 1) %>%
  pull(iteration)

history_df %>%
  ggplot(aes(x = iteration, y = loss)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_vline(xintercept = steepest, linetype = "dashed", color = "firebrick") +
  annotate("text",
           x     = steepest + 25,
           y     = max(history_df$loss) * 0.92,
           label = "Steepest\ndescent",
           color = "firebrick", size = 3.5, hjust = 0) +
  labs(
    title = "Cross-entropy loss over training iterations",
    x     = "Iteration",
    y     = "Loss"
  ) +
  theme_minimal()
```

---

## Visualization 3: Gradient Magnitude by Layer

Backpropagation pushes gradient signals backward from the output layer toward the input layer. Here we track the  norm of each layer's gradient at every iteration. If gradients in W1 (closer to the input) are consistently smaller than those in W2, that foreshadows the **vanishing gradient** problem common in deeper networks.

```{r viz-gradients}
history_df %>%
  pivot_longer(c(dW1_norm, dW2_norm),
               names_to  = "layer",
               values_to = "grad_norm") %>%
  mutate(layer = recode(layer,
                        dW1_norm = "W1 (input → hidden)",
                        dW2_norm = "W2 (hidden → output)")) %>%
  ggplot(aes(x = iteration, y = grad_norm, color = layer)) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = c(
    "W1 (input → hidden)"    = "steelblue",
    "W2 (hidden → output)"   = "firebrick"
  )) +
  labs(
    title = "Gradient magnitude by layer over training",
    x     = "Iteration",
    y     = "Norm of weight update",
    color = "Layer"
  ) +
  theme_minimal()
```

---

## Visualization 4: W2 Weight Heatmap — Before vs. After Training

Backpropagation doesn't update weights randomly — it nudges them in directions that reduce loss. The W2 matrix (hidden → output, 10 × 5) is small enough to visualize directly. Compare the random noise of initialization to the structured pattern that emerges after training.

```{r viz-w2-heatmap}
tidy_W2 <- function(net, label) {
  net$W2 %>%
    as_tibble(.name_repair = "minimal") %>%
    set_names(str_c("hidden_", seq_len(ncol(.)))) %>%
    mutate(output_class = row_number(), phase = label) %>%
    pivot_longer(-c(output_class, phase),
                 names_to  = "hidden_unit",
                 values_to = "weight")
}

bind_rows(
  tidy_W2(network_init,    "Before training"),
  tidy_W2(network_trained, "After training")
) %>%
  mutate(phase = fct_inorder(phase)) %>%
  ggplot(aes(x = hidden_unit, y = factor(output_class), fill = weight)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "firebrick",
                       midpoint = 0) +
  facet_wrap(~phase) +
  labs(
    title = "W2 weights: before vs. after training",
    x     = "Hidden unit",
    y     = "Output class",
    fill  = "Weight"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## Visualization 5: Softmax Output — One Sample, Before vs. After Training

The forward pass ends with a softmax layer that converts raw scores into a probability distribution over classes. Before training, probability is spread roughly uniformly. After training, it should concentrate on the correct class.

```{r viz-softmax}
forward_probs <- function(net, x) {
  z1 <- net$W1 %*% x + net$b1
  a1 <- pmax(0, z1)
  z2 <- net$W2 %*% a1 + net$b2
  ez <- exp(z2 - max(z2))
  as.vector(ez / sum(ez))
}

sample_idx <- 1
x_sample   <- synth$X[sample_idx, ]
true_class  <- which(synth$Y[sample_idx, ] == 1)

bind_rows(
  tibble(class = 1:j,
         prob  = forward_probs(network_init, x_sample),
         phase = "Before training"),
  tibble(class = 1:j,
         prob  = forward_probs(network_trained, x_sample),
         phase = "After training")
) %>%
  mutate(
    phase   = fct_inorder(phase),
    correct = class == true_class
  ) %>%
  ggplot(aes(x = factor(class), y = prob, fill = correct)) +
  geom_col() +
  scale_fill_manual(values = c("FALSE" = "grey75", "TRUE" = "firebrick"),
                    labels  = c("Other class", "True class")) +
  facet_wrap(~phase) +
  labs(
    title    = "Softmax output for one sample: before vs. after training",
    subtitle = paste("True class:", true_class),
    x        = "Class",
    y        = "Predicted probability",
    fill     = NULL
  ) +
  theme_minimal()
```

---

## Visualization 6: ReLU Activation Sparsity

The ReLU activation function outputs zero whenever its input is negative, effectively "silencing" that hidden unit for a given sample. Here we look at the fraction of hidden units that are silent (= 0) for each sample after training, broken out by true class. High sparsity means most neurons are inactive — which can be either a sign of efficient coding or a warning of dead neurons.

```{r viz-relu-sparsity}
relu_sparsity <- function(net, X) {
  Z1 <- X %*% t(net$W1) +
    matrix(rep(net$b1, each = nrow(X)), nrow = nrow(X))
  A1 <- pmax(Z1,0)
  rowMeans(A1 == 0)
}

tibble(
  sparsity = relu_sparsity(network_trained, synth$X),
  class    = factor(apply(synth$Y, 1, which.max))
) %>%
  ggplot(aes(x = sparsity, fill = class)) +
  geom_histogram(bins = 10, position = "identity", alpha = 0.5) +
  labs(
    title = "ReLU sparsity in the hidden layer after training",
    x     = "Fraction of hidden units = 0 (silenced by ReLU)",
    y     = "Count",
    fill  = "True class"
  ) +
  theme_minimal()
```
