---
title: "Lecture 18: Multinomial Classification with Keras and ELS:2002"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.width = 7,
  fig.height = 5
)

# Set to FALSE after first run to skip expensive computations
fit_model <- FALSE
```

## Overview

In the previous lectures we built intuition for neural networks, worked through
backpropagation by hand, and explored regularization strategies. Today we put
all of that together in a real applied problem: predicting college major field of
study from the Education Longitudinal Study of 2002 (ELS:2002), using a
multilayer perceptron (MLP) fit via **Keras** inside the tidymodels framework.

Along the way we will pay particular attention to the features Keras provides
that you either cannot do at all in PyTorch, or that require substantially more
boilerplate. The headline items:

1. **Callbacks** — early stopping, learning rate scheduling, and custom logging
   are first-class objects in Keras, wired directly into `model.fit()`.
2. **`validation_split`** — Keras can carve out a held-out slice of training
   data automatically, with no extra code from you.
3. **Opinionated defaults** — Keras ships with sensible defaults for optimizers,
   loss functions, and initializers. For a multinomial problem this means you
   get `softmax` output and `categorical_crossentropy` for free.
4. **Training history** — Keras returns a `History` object after every fit,
   giving you epoch-level loss and metric curves without writing a custom loop.

PyTorch is more flexible and often faster, but that flexibility comes at the
cost of writing more infrastructure. Keras trades some of that ceiling for a
much lower floor. For applied social science and education research — where the
bottleneck is usually data and domain knowledge, not model architecture — that
tradeoff is often worth it.

---

## Setup and Libraries

```{r libraries}
library(tidyverse)
library(haven)
library(here)
library(janitor)
library(tidymodels)
library(keras)      # Keras R interface (wraps TensorFlow)
library(vip)        # Permutation-based variable importance
library(kableExtra)
```

```{r theme}
# Shared plot theme used throughout
theme_els <- function() {
  theme_minimal(base_size = 11) +
    theme(
      plot.title    = element_text(face = "bold", size = 12),
      plot.subtitle = element_text(color = "grey40", size = 10),
      axis.title    = element_text(size = 10),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
}
```

---

## The Research Problem

ELS:2002 followed a nationally representative cohort of 10th graders through
high school and into postsecondary education. Our analytic sample consists of
students who:

- Ever enrolled in postsecondary education (`f2evratt == 1`), and
- Completed a bachelor's degree with an observed CIP-2 major code
  (`f3tzbch1cip2` is not missing).

The outcome is **collapsed CIP-2 major field**, a multinomial variable with
several broad categories (e.g., STEM, Education, Business, Humanities). We
predict it from a set of longitudinal predictors spanning 10th grade background,
12th grade attitudes and plans, and early postsecondary transcript data —
carefully ordered to avoid data leakage. Declared major is excluded as a
predictor for the same reason.

Race and gender are withheld from the predictor set entirely; they are reserved
for a post-hoc subgroup analysis.

---

## Data Wrangling

### Variable Inventory

The variable selection spans four ELS waves. Broadly:

- **BY (base year, 10th grade):** demographics, cognitive achievement scores,
  psychosocial scales, aspirations, school context.
- **F1 (12th grade):** updated expectations, highest math course, PS plans.
- **F2 (2006):** first postsecondary institution characteristics, enrollment
  behavior.
- **F3TZ (postsecondary transcript, early years only):** credits earned and GPA
  in years 1–2. *Only early-enrollment transcript variables are used* — later
  transcript variables would leak information about the major itself.

```{r els-vars}
els_vars <- c(
  "stu_id",

  # Outcome
  "f3tzbch1cip2",

  # Student background (BY)
  "bygnstat", "bystlang", "byfcomp", "bysibhom",
  "bypared", "bymothed", "byfathed", "bygpared",
  "byoccum", "byoccuf", "byincome", "byses2",
  "bygrdrpt", "byriskfc",

  # Cognitive achievement (BY)
  "bytxmstd", "bytxrstd",

  # Psychosocial scales (BY)
  "bymathse", "byenglse", "byconexp", "byinstmo",
  "byactctl", "bystprep", "byacclim",

  # Aspirations & plans (BY)
  "bystexp", "byparasp", "byocc30",

  # Behavioral engagement (BY)
  "byhmwrk", "bynsport", "byxtracu", "byworksy", "bywrkhrs",

  # School context (BY)
  "bysctrl", "byurban", "byregion",

  # F1 wave (senior year 2004)
  "f1stexp", "f1byedex", "f1mathse",
  "f1occ30", "f1psepln",
  "f1ses2",  "f1xtracu", "f1wrkhrs",
  "f1colinf", "f1himath", "f1pared",

  # F2 postsecondary context (2006)
  "f2evratt",  "f2ps1lvl", "f2ps1sec",
  "f2ps1ftp",  "f2ps1rem", "f2ps1aid",
  "f2psstrt",  "f2enrgap", "f2switch",
  "f2stexp",   "f2f1edex", "f2pseexm", "f2psepln",
  "f2psapsl",  "f2ptn3ps",

  # F3TZ transcript — early enrollment only (years 1–2; safe predictors)
  "f3tzps1sec", "f3tzps1slc",
  "f3tzhs2ps1",
  "f3tzyr1ern", "f3tzyr2ern", "f3tzyr12ern",
  "f3tzyr1gpa", "f3tzyr2gpa",
  "f3tzremtot", "f3tzremmttot", "f3tzrementot",
  "f3tznumcrswd",

  # Subgroup analysis variables — NOT model predictors
  "bysex", "byrace"
)

# Nominal variables to cast as factors before the recipe
nominal_vars <- c(
  "bygnstat", "bystlang", "byfcomp",
  "byoccum",  "byoccuf",  "byocc30",
  "bysctrl",  "byurban",  "byregion",
  "byworksy",
  "f1psepln", "f1colinf", "f1himath", "f1occ30",
  "f2ps1lvl", "f2ps1sec",
  "f2ps1ftp", "f2ps1rem", "f2ps1aid",
  "f2enrgap", "f2switch", "f2pseexm",
  "f2ptn3ps",
  "f3tzps1sec", "f3tzps1slc"
)
```

### Loading and Cleaning

NCES uses negative codes (–1 through –9) to encode various forms of
non-response and skip patterns. We recode all of these to `NA` after loading.

```{r load-data}
load_els <- function(path) {
  read_dta(path) %>%
    clean_names() %>%
    select(any_of(els_vars)) %>%
    mutate(across(everything(), haven::zap_labels)) %>%
    mutate(across(where(is.numeric), ~ replace(.x, .x %in% -9:-1, NA)))
}

load_cip_labels <- function(path) {
  read_csv(path, show_col_types = FALSE)
}

wrangle_els <- function(els_raw, cip_labels) {
  cip_collapse <- cip_labels %>%
    select(cip2_code, cip2_collapsed_code, cip2_collapsed_label) %>%
    distinct()

  els_raw %>%
    mutate(across(any_of(nominal_vars), as.factor)) %>%
    left_join(cip_collapse, by = join_by(f3tzbch1cip2 == cip2_code)) %>%
    mutate(
      cip2 = factor(
        cip2_collapsed_code,
        levels = sort(unique(cip_collapse$cip2_collapsed_code))
      )
    )
}

make_analytic_sample <- function(els_clean) {
  els_clean %>%
    filter(f2evratt == 1, !is.na(cip2))
}
```

```{r run-wrangle, eval=fit_model}
els_raw       <- load_els(here("scripts","els_02_12_byf3pststu_v1_0.dta"))
cip_labels    <- load_cip_labels(here("scripts","cip2_labels.csv"))
els_clean     <- wrangle_els(els_raw, cip_labels)
analytic_sample <- make_analytic_sample(els_clean)

saveRDS(analytic_sample, here("analytic_sample.rds"))
```

```{r load-analytic-sample, eval=!fit_model}
analytic_sample <- readRDS(here("scripts","analytic_sample.rds"))
cip_labels      <- read_csv(here("scripts","cip2_labels.csv"),
                                   show_col_types = FALSE)
```

### Outcome Distribution

```{r outcome-dist, fig.height=6}
label_lookup <- cip_labels %>%
  select(cip2_collapsed_code, cip2_collapsed_label) %>%
  distinct() %>%
  mutate(cip2_collapsed_code = as.factor(cip2_collapsed_code))

analytic_sample %>%
  count(cip2) %>%
  mutate(pct = n / sum(n)) %>%
  left_join(label_lookup, by = join_by(cip2 == cip2_collapsed_code)) %>%
  mutate(cip2_collapsed_label = fct_reorder(cip2_collapsed_label, pct)) %>%
  ggplot(aes(x = pct, y = cip2_collapsed_label)) +
  geom_col(fill = "#2C7BB6", alpha = 0.85) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title    = "Distribution of Bachelor's Degree Major Fields",
    subtitle = "Analytic sample: ELS:2002 bachelor's degree completers",
    x        = "Percent of sample",
    y        = NULL
  ) +
  theme_els()
```

The outcome is imbalanced — Business and Social Sciences are far more common
than Computer Science or Engineering. This has direct implications for how we
choose metrics (more on that in the evaluation section) and how Keras handles
the output layer.

### Missingness

```{r missingness-plot, fig.height=8}
analytic_sample %>%
  select(-stu_id, -bysex, -byrace, -f3tzbch1cip2,
         -cip2_collapsed_code, -cip2_collapsed_label) %>%
  summarise(across(everything(),
                   list(n_missing   = ~ sum(is.na(.x)),
                        pct_missing = ~ mean(is.na(.x))))) %>%
  pivot_longer(everything(),
               names_to    = c("variable", ".value"),
               names_pattern = "^(.+)_(n_missing|pct_missing)$") %>%
  filter(pct_missing > 0) %>%
  mutate(variable = fct_reorder(variable, pct_missing)) %>%
  ggplot(aes(x = pct_missing, y = variable)) +
  geom_col(fill = "#D7191C", alpha = 0.75) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title    = "Missing Data by Predictor Variable",
    subtitle = "Analytic sample prior to imputation",
    x        = "Percent missing",
    y        = NULL
  ) +
  theme_els()
```

Transcript variables (`f3tzyr*`) show the highest missingness — students who
did not attend college, attended briefly, or whose records were not matched have
missing values here. Crucially, *the fact of missingness is itself informative*:
a student with no Year 1 GPA record is different from one with a 2.5. We encode
this with missingness indicators in the recipe.

---

## Train/Test Split and Cross-Validation

```{r split}
set.seed(2002)
els_split <- initial_split(analytic_sample, prop = 0.80, strata = cip2)
els_train <- training(els_split)
els_test  <- testing(els_split)

```

We use stratified splits to preserve the class distribution across training and
test sets — important given the imbalance we saw above.

```{r folds, eval=fit_model}
set.seed(2002)
els_folds <- vfold_cv(els_train, v = 5, strata = cip2)
```

---

## Preprocessing Recipe

The recipe codifies every preprocessing step in a reproducible, leak-free way.
Steps are applied in order, and all parameters (medians, standard deviations)
are estimated only from the training fold — never from the test data.

```{r recipe}
els_recipe <- recipe(cip2 ~ ., data = els_train) %>%

  # Non-predictor roles — included in the data but excluded from the model matrix
  update_role(stu_id,              new_role = "id") %>%
  update_role(bysex,               new_role = "subgroup") %>%
  update_role(byrace,              new_role = "subgroup") %>%
  update_role(f3tzbch1cip2,        new_role = "original_outcome") %>%
  update_role(cip2_collapsed_code,  new_role = "id") %>%
  update_role(cip2_collapsed_label, new_role = "id") %>%

  # Missingness indicators before imputation
  # The *fact* of missing transcript data carries signal about PS pathways
  step_indicate_na(f3tzyr1ern, f3tzyr2ern, f3tzyr1gpa, f3tzyr2gpa) %>%

  # Imputation
  step_impute_median(all_numeric_predictors()) %>%
  step_impute_mode(all_nominal_predictors()) %>%

  # Dummy encoding
  # step_novel() guards against factor levels in CV/test folds not seen in training
  step_novel(all_nominal_predictors()) %>%
  # one_hot = TRUE: no reference category dropped; Keras needs dense numeric input
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%

  # Housekeeping
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())
```

**Why `one_hot = TRUE`?** Standard dummy encoding drops one reference category
per factor to avoid the dummy variable trap in linear models. Neural networks
have no such constraint — no matrix inversion, no multicollinearity problem —
and symmetric encoding can help the network learn representations for all levels
equally. We keep all dummies.

**Why normalize?** Keras uses gradient-based optimization (Adam by default).
If predictors are on wildly different scales, gradients will be dominated by
large-magnitude variables, and training will be slow or unstable. Normalizing to
mean 0, standard deviation 1 is not optional here — it is required.

```{r check-recipe}
# Face validity: how many predictors after preprocessing?
prepped <- prep(els_recipe, training = els_train)
baked   <- bake(prepped, new_data = NULL)

cat("Rows in baked training data:", nrow(baked), "\n")
cat("Columns (outcome + predictors):", ncol(baked), "\n")
cat("Number of predictors:", ncol(baked) - 1, "\n")
```

---

## Keras-Specific Features

Before we build the model spec, it's worth pausing to understand what Keras is
doing under the hood and where it differs from PyTorch.

### The Architecture

We use a two-hidden-layer MLP (the default for `parsnip`'s Keras engine). Each
hidden layer uses **ReLU** activation. The output layer uses **softmax**, which
converts the raw logit scores into a proper probability distribution over all
major categories. This is automatic in Keras — you do not specify the output
layer manually when using `parsnip`.

```
Input (p predictors)
    ↓
Hidden Layer 1: hidden_units nodes, ReLU + Dropout
    ↓
Hidden Layer 2: hidden_units nodes, ReLU + Dropout
    ↓
Output: K nodes (one per major category), Softmax
```

### Feature 1: Callbacks

Keras **callbacks** are objects that can inspect and modify training at the end
of each epoch. PyTorch has no built-in equivalent — you write a training loop
and add your own logic. Keras callbacks are first-class objects that plug
directly into `model.fit()`.

The most important for us is **early stopping**: we specify a maximum number of
epochs, but allow training to halt early if the validation loss stops improving.

```{r early-stopping-demo, eval=FALSE}
# This is what Keras runs internally when you pass callbacks via set_engine()
callback_early_stopping(
  monitor              = "val_loss",   # watch held-out validation loss
  patience             = 15,           # tolerate 15 epochs without improvement
  restore_best_weights = TRUE          # rewind to the best epoch's weights
)
```

`restore_best_weights = TRUE` is subtle but important: without it, you get the
weights from the *last* epoch, which may be worse than an earlier checkpoint.
Keras handles this bookkeeping for you. In PyTorch you would save a checkpoint
manually at each epoch and reload it at the end.

Other useful callbacks you can pass the same way:

```{r other-callbacks, eval=FALSE}
# Reduce learning rate when improvement plateaus (learning rate scheduling)
callback_reduce_lr_on_plateau(
  monitor  = "val_loss",
  factor   = 0.5,          # multiply LR by 0.5 when plateau detected
  patience = 5
)

# Write TensorBoard logs for interactive visualization
callback_tensorboard(log_dir = "logs/run_01")

# Save model weights after each epoch (model checkpointing)
callback_model_checkpoint(
  filepath          = "checkpoints/weights_{epoch}.keras",
  save_best_only    = TRUE
)
```

You can combine multiple callbacks in a list — Keras runs all of them in
sequence at the end of each epoch. All of this is in Keras's standard library.
PyTorch users typically reach for separate packages (`lightning`, `ignite`) or
write this infrastructure themselves.

### Feature 2: `validation_split`

When you pass `validation_split = 0.2` to `set_engine("keras", ...)`, Keras
automatically holds out the last 20% of the training data (by row order) as a
validation set. This validation set is used by callbacks to monitor performance
— it is not the test set.

In PyTorch you would split the `DataLoader` manually, or use a
`SubsetRandomSampler`. Keras does it in one argument.

```{r note-validation-split, echo=FALSE}
# Note: The validation_split used inside Keras training is separate from the
# tidymodels cross-validation folds. CV folds evaluate hyperparameter
# combinations; the internal validation_split gives early stopping its signal
# within each fold's training run.
```

> **Important distinction:** The 20% `validation_split` inside Keras is *not*
> the same as the tidymodels CV folds. The CV folds evaluate which
> hyperparameter combination is best. The internal `validation_split` gives the
> early stopping callback its signal within each individual training run. They
> are nested: for each fold/grid-point combination, Keras trains on 80% of that
> fold's training data and monitors loss on the other 20%.

### Feature 3: Training History

Every Keras `model.fit()` call returns a `History` object with epoch-by-epoch
loss and metric values for both training and validation. Through the parsnip
interface this is less directly accessible, but when you use Keras directly you
get it for free:

```{r history-demo, eval=FALSE}
# Direct Keras API (outside of tidymodels)
history <- model %>% fit(
  x_train, y_train,
  epochs           = 200,
  validation_split = 0.2,
  callbacks        = list(callback_early_stopping(patience = 15))
)

# Plot training vs. validation loss over epochs — built in
plot(history)
```

In PyTorch you append to a list inside your training loop. Keras returns it
automatically.

---

## Model Specification

```{r model-spec}
keras_spec <- mlp(
  hidden_units = tune(),
  epochs       = tune(),
  penalty      = 0,       # fixed — cannot specify alongside dropout in Keras engine
  dropout      = tune()
  # learn_rate is not exposed by the Keras parsnip engine
  # Adam optimizer uses its default (0.001)
) %>%
  set_engine(
    "keras",
    validation_split = 0.2,
    callbacks = list(
      keras::callback_early_stopping(
        monitor              = "val_loss",
        patience             = 15,
        restore_best_weights = TRUE
      )
    )
  ) %>%
  set_mode("classification")
```

**The `penalty`/`dropout` constraint.** The Keras parsnip engine does not allow
`penalty` (L2 regularization) and `dropout` to be tuned simultaneously in a
single `mlp()` call. This is an engine-level limitation, not a Keras limitation.
If you need both, the cleanest solution is to run two separate tuning workflows
(dropout-only and L2-only) and compare their best configurations. We focus on
dropout here.

**The `learn_rate` situation.** The Keras parsnip engine does not expose
`learn_rate` as a tunable parameter through `mlp()`. Adam's default (0.001) is
used. If you need to tune the learning rate, you can pass a custom optimizer
through engine arguments:

```{r lr-engine-arg, eval=FALSE}
# Advanced: custom learning rate via engine argument
mlp(...) %>%
  set_engine(
    "keras",
    optimizer = keras::optimizer_adam(learning_rate = 0.0005)
  )
```

---

## Workflow

```{r workflow}
els_workflow <- workflow() %>%
  add_recipe(els_recipe) %>%
  add_model(keras_spec)
```

---

## Hyperparameter Tuning

### Tuning Grid

We'll use a space-filling Latin hypercube design. Rather than an exhaustive
grid (which scales as $k^p$ where $k$ is levels and $p$ is parameters), a
space-filling design spreads points evenly across the hyperparameter space,
giving much better coverage with far fewer evaluations.

```{r tuning-grid}
set.seed(2002)
tuning_grid <- grid_space_filling(
  hidden_units(range = c(32L, 256L)),
  epochs(range       = c(50L, 300L)),
  dropout(range      = c(0, 0.5)),
  size = 20
)

tuning_grid
```

### Running the Tuning

```{r tune-model, eval=fit_model}
# Keras manages its own internal threading via TensorFlow.
# Running multiple Keras sessions in parallel R workers causes memory conflicts
# and is typically slower on CPU. We keep allow_par = FALSE.
tuning_results <- tune_grid(
  els_workflow,
  resamples = els_folds,
  grid      = tuning_grid,
  metrics   = metric_set(roc_auc, accuracy, mn_log_loss),
  control   = control_grid(
    save_pred = TRUE,
    verbose   = TRUE,
    allow_par = FALSE
  )
)

saveRDS(tuning_results, here("scripts","tuning_results.rds"))
```

```{r load-tuning-results, eval=!fit_model}
tuning_results <- readRDS(here("scripts","tuning_results.rds"))
```

### Examining Tuning Results

```{r tuning-overview}
# Summary across all grid points
collect_metrics(tuning_results, summarize = TRUE) %>%
  filter(.metric == "roc_auc") %>%
  arrange(desc(mean)) %>%
  select(hidden_units, epochs, dropout, mean, std_err) %>%
  head(5) %>%
  kable(digits = 3, caption = "Top 5 hyperparameter combinations by ROC AUC") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

```{r tuning-plot, fig.height=5}
# Ranked performance plot with dropout color-coded
collect_metrics(tuning_results) %>%
  filter(.metric == "roc_auc") %>%
  arrange(desc(mean)) %>%
  mutate(rank = row_number()) %>%
  ggplot(aes(x = rank, y = mean)) +
  geom_point(aes(color = dropout), size = 3, alpha = 0.8) +
  geom_errorbar(aes(ymin = mean - std_err, ymax = mean + std_err),
                width = 0.3, alpha = 0.4) +
  scale_color_viridis_c(option = "plasma") +
  labs(
    title    = "Hyperparameter Tuning Results",
    subtitle = "ROC AUC (macro-weighted) across Latin hypercube grid, 5-fold CV",
    x        = "Rank (best to worst)",
    y        = "Mean ROC AUC",
    color    = "Dropout rate"
  ) +
  theme_els()
```

```{r tuning-marginals, fig.height=4}
# How does each parameter relate to performance?
collect_metrics(tuning_results) %>%
  filter(.metric == "roc_auc") %>%
  select(mean, hidden_units, epochs, dropout) %>%
  pivot_longer(c(hidden_units, epochs, dropout),
               names_to = "parameter", values_to = "value") %>%
  ggplot(aes(x = value, y = mean)) +
  geom_point(alpha = 0.7, color = "#2C7BB6") +
  geom_smooth(method = "loess", se = FALSE, color = "#D7191C") +
  facet_wrap(~ parameter, scales = "free_x") +
  labs(
    title = "ROC AUC vs. Each Hyperparameter (Marginal)",
    subtitle = "Loess smoother — other parameters vary freely",
    y = "Mean ROC AUC"
  ) +
  theme_els()
```

```{r face-validity}
# Face validity: compare best model to null classifier
null_acc <- analytic_sample %>%
  count(cip2) %>%
  mutate(pct = n / sum(n)) %>%
  slice_max(pct, n = 1) %>%
  pull(pct)

best_roc <- show_best(tuning_results, metric = "roc_auc", n = 1) %>%
  pull(mean)

cat("Null classifier accuracy (always predict most common class):",
    round(null_acc, 3), "\n")
cat("Best model ROC AUC (macro-weighted):                         ",
    round(best_roc, 3), "\n")
```

```{r tuning-range}
# Is there meaningful variation across grid points?
collect_metrics(tuning_results) %>%
  filter(.metric == "roc_auc") %>%
  summarise(
    min   = min(mean),
    max   = max(mean),
    range = max(mean) - min(mean),
    sd    = sd(mean)
  ) %>%
  kable(digits = 3, caption = "Performance range across tuning grid") %>%
  kable_styling(bootstrap_options = "condensed")
```

## Selecting Best Parameters and Final Fit

```{r select-best}
best_params <- select_best(tuning_results, metric = "roc_auc")
best_params

# Finalize the workflow with the best hyperparameter values
final_workflow <- finalize_workflow(els_workflow, best_params)
```

`last_fit()` fits the finalized model on the **full training set** and evaluates
it on the held-out **test set**. This is the only time the test set is touched.

```{r final-fit}
set.seed(2002)
final_fit <- last_fit(
  final_workflow,
  split   = els_split,
  metrics = metric_set(roc_auc, accuracy, mn_log_loss)
)

# Save fit for prediction later-- some weirdness with keras
fitted_wf <- extract_workflow(final_fit)

#saveRDS(final_fit, here("scripts","final_fit.rds"))
```

```{r load-final-fit, eval=!fit_model}
#final_fit <- readRDS(here("scripts","final_fit.rds"))
```


## Evaluation

### Why ROC AUC for an Imbalanced Multinomial Outcome?

Accuracy is a poor metric here. If Business majors make up 25% of the sample,
a classifier that always predicts "Business" achieves 25% accuracy — not
meaningfully worse than a model that actually learned something. ROC AUC
sidesteps this by evaluating the model's ability to *rank* classes, not just
assign the most common one. We use **macro-weighted** AUC, which averages
one-vs-rest AUC scores weighted by class frequency.

Multinomial log-loss (`mn_log_loss`) is also reported; it penalizes confident
wrong predictions heavily, which is useful for calibration assessment.

### Overall Metrics

```{r overall-metrics}
collect_metrics(final_fit) %>%
  select(.metric, .estimate) %>%
  rename(metric = .metric, estimate = .estimate) %>%
  kable(digits = 3, caption = "Test set performance metrics") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

### Test Set Predictions

```{r test-predictions}
test_preds <- collect_predictions(final_fit) %>%
  bind_cols(
    els_test %>% select(stu_id, bysex, byrace, cip2_collapsed_label)
  )

glimpse(test_preds)
```

### Confusion Matrix

```{r confusion-matrix, fig.width=10, fig.height=9}
# Build a tidy confusion matrix with human-readable major labels on both axes.
# autoplot() only shows the numeric cip2 factor levels, so we do this manually.
#
# Row-normalise so the diagonal reads as recall per class — more informative
# than raw counts when class sizes differ substantially.

all_codes <- levels(test_preds$cip2)

ordered_labels <- label_lookup %>%
  filter(as.character(cip2_collapsed_code) %in% all_codes) %>%
  arrange(match(as.character(cip2_collapsed_code), all_codes)) %>%
  pull(cip2_collapsed_label) %>%
  str_wrap(width = 22)

test_preds %>%
  count(truth = cip2, predicted = .pred_class, .drop = FALSE) %>%
  group_by(truth) %>%
  mutate(row_pct = n / sum(n)) %>%
  ungroup() %>%
  left_join(
    label_lookup %>%
      mutate(truth_label = str_wrap(cip2_collapsed_label, width = 22)) %>%
      select(cip2_collapsed_code, truth_label),
    by = join_by(truth == cip2_collapsed_code)
  ) %>%
  left_join(
    label_lookup %>%
      mutate(pred_label = str_wrap(cip2_collapsed_label, width = 22)) %>%
      select(cip2_collapsed_code, pred_label),
    by = join_by(predicted == cip2_collapsed_code)
  ) %>%
  mutate(
    truth_label = factor(truth_label, levels = ordered_labels),
    pred_label  = factor(pred_label,  levels = ordered_labels)
  ) %>%
  ggplot(aes(x = pred_label, y = fct_rev(truth_label), fill = row_pct)) +
  geom_tile(color = "white", linewidth = 0.35) +
  geom_text(
    aes(label = ifelse(n > 0, as.character(n), "")),
    size = 2.5, color = "grey20"
  ) +
  scale_fill_gradient(
    low    = "white",
    high   = "#2C7BB6",
    limits = c(0, 1),
    labels = scales::percent_format(accuracy = 1),
    name   = "Row %\n(recall)"
  ) +
  scale_x_discrete(position = "top") +
  labs(
     title    = "Confusion Matrix — Test Set Predictions",
    subtitle = "Rows: true class  |  Columns: predicted class",
    x        = NULL,      # <-- remove x title, it's creating the gap
    y        = "True major"
  ) +
   coord_cartesian(clip = "off") +
    theme_els() +
  theme(
    axis.text.x      = element_text(angle = -40, hjust = 1, size = 7),
    axis.text.y      = element_text(size = 7),
    legend.position  = "right",
    legend.key.width = unit(0.4, "cm"),
    plot.margin      = margin(t = 2, r = 5, b = 5, l = 5, unit = "mm")  # <-- tighten top
  )
```


### Per-Class Precision, Recall, and F1

```{r per-class-metrics}
per_class <- bind_rows(
  precision(test_preds, truth = cip2, estimate = .pred_class,
            estimator = "macro") %>% mutate(metric = "precision"),
  recall(test_preds, truth = cip2, estimate = .pred_class,
         estimator = "macro") %>% mutate(metric = "recall"),
  f_meas(test_preds, truth = cip2, estimate = .pred_class,
         estimator = "macro") %>% mutate(metric = "f1")
)

per_class %>%
  select(metric, .estimate) %>%
  kable(digits = 3, caption = "Macro-averaged per-class metrics") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

---

## Variable Importance

Keras models do not have interpretable coefficients — there is no `coef()`
method. This is the tradeoff for the nonlinearity. We recover feature importance
via **permutation**: for each predictor, we shuffle its values across all
observations, generate new predictions, and measure how much the model's
macro-weighted ROC AUC drops. A large drop means the model relied heavily on
that feature.

**Permutation importance** is the standard approach for interpreting models — like neural networks — that have no coefficients to read off directly.

The core idea is to ask: *how much does the model's performance degrade when it can no longer use a particular feature?* We operationalize this by destroying the feature's information content without changing the shape of the data. Rather than removing the column entirely (which would break the model's input dimensions), we shuffle its values randomly across all observations. This severs the relationship between that feature and the outcome while leaving everything else intact.

Concretely, for each predictor we:

1. Start with a **baseline** — run the fitted model on a held-out sample and record its macro-weighted ROC AUC. This is the model's performance when it has full access to all features.

2. **Permute** — randomly reorder the values of one predictor column, so that each student now has some other student's value for that variable. The model still sees a number in that column; it just no longer carries real information about that student.

3. **Re-score** — run the model again on the permuted data and record the new AUC.

4. **Importance = baseline AUC − permuted AUC.** If the feature mattered, the model's ability to rank students by their true major deteriorates noticeably. If the feature was irrelevant or redundant, the AUC barely changes.

We repeat steps 2–4 five times (`nsim = 5`) for each feature and average the drops, because any single shuffle is one random draw — averaging over several gives a more stable estimate.

**Why ROC AUC specifically?** Because our outcome is multinomial and imbalanced. Accuracy would be dominated by the most common major categories. Macro-weighted AUC evaluates the model's ability to discriminate each class from all others, weighted by class frequency, so every major field contributes to the score roughly in proportion to how many students are in it.

**Why a 75% sample rather than the full training set?** Purely for speed — we are running hundreds of predict calls (one per feature per simulation). The 75% sample is drawn once and held fixed across all permutations so that the baseline and every permuted AUC are computed on exactly the same rows, making the differences comparable.

**What the numbers mean:** An importance of 0.03 means permuting that feature caused macro-weighted AUC to drop by about 3 percentage points on average. A feature near zero had essentially no effect when scrambled — either the model wasn't using it, or its information was fully recoverable from correlated predictors that remained intact.

`vip::vi_permute()` does not support multiclass probability metrics — its
`metric` argument passes vectors to a scalar function, but `roc_auc_vec()` for
multiclass requires a full probability matrix. We therefore implement the loop
directly, which is only slightly more code and makes the algorithm completely
transparent.

The logic is:

1. Get a baseline AUC on a 75% sample of the training data.
2. For each predictor column, shuffle its values (breaking the predictor-outcome
   association), predict, compute AUC, and record the drop from baseline.
3. Repeat `nsim` times and average the drops.

```{r vip-helpers}
# Compute macro-weighted ROC AUC from a fitted workflow and a data frame.
# Returns a single scalar.
compute_auc <- function(fitted_wf, data) {
  probs <- predict(fitted_wf, data, type = "prob")
  truth <- data[["cip2"]]

  bind_cols(probs, tibble(cip2 = truth)) %>%
    roc_auc(
      truth     = cip2,
      matches("^\\.pred_"),       # <-- regex match works inside functions
      estimator = "macro_weighted"
    ) %>%
    pull(.estimate)
}
# Permutation importance for a single predictor column.
# Returns the mean AUC drop across nsim permutations.
permute_one <- function(fitted_wf, data, col, nsim) {
  map_dbl(seq_len(nsim), function(i) {
    permuted <- data %>%
      mutate(across(all_of(col), sample))
    baseline_auc - compute_auc(fitted_wf, permuted)
  }) %>%
    mean()
}
```


```{r vip}
# Pull the raw prob matrix once per permutation at the top level,
# rather than calling predict() inside a helper function.

set.seed(2002)
vip_sample <- els_train %>% slice_sample(prop = 0.75)

# Baseline: predict once at top level, score it
baseline_probs <- augment(fitted_wf, vip_sample)

score_auc <- function(df) {
  df %>%
    roc_auc(
      truth     = cip2,
      matches("^\\.pred_(?!class)",perl=TRUE),
      estimator = "macro_weighted"
    ) %>%
    pull(.estimate)
}

baseline_auc <- score_auc(baseline_probs)
cat("Baseline AUC:", round(baseline_auc, 4), "\n")

pred_cols <- names(baseline_probs) %>% keep(~ startsWith(.x, ".pred_"))
non_pred_cols <- c(
  "cip2", "stu_id", "bysex", "byrace",
  "f3tzbch1cip2", "cip2_collapsed_code", "cip2_collapsed_label",
  ".pred_class", pred_cols
)
pred_features <- setdiff(names(vip_sample), non_pred_cols)

# For each feature: permute in vip_sample, re-augment at top level, score
permute_one <- function(col, nsim = 5) {
  map_dbl(seq_len(nsim), function(i) {
    permuted_data <- vip_sample %>%
      mutate(across(all_of(col), sample))
    permuted_probs <- augment(fitted_wf, permuted_data)
    baseline_auc - score_auc(permuted_probs)
  }) %>% mean()
}

vip_scores <- tibble(Variable = pred_features) %>%
  mutate(Importance = map_dbl(Variable, permute_one)) %>%
  arrange(desc(Importance))

write_rds(vip_scores,file=here("scripts","vip_scores.rds"))
```



```{r load-vip, eval=!fit_model}
vip_scores <- readRDS(here("scripts","vip_scores.rds"))
```

```{r vip-plot, fig.height=8}
vip_scores %>%
  slice_max(Importance, n = 30) %>%
  mutate(Variable = fct_reorder(Variable, Importance)) %>%
  ggplot(aes(x = Importance, y = Variable)) +
  geom_col(fill = "#2C7BB6", alpha = 0.85) +
  labs(
    title    = "Permutation-Based Variable Importance (Top 30)",
    subtitle = "Mean decrease in macro-weighted ROC AUC from permuting each feature (5 sims, 75% sample)",
    x        = "Mean AUC drop",
    y        = NULL
  ) +
  theme_els()
```

**Caveats to keep in mind:**

- Correlated features *understate* individual importance — if `bytxmstd` and
  `bytxrstd` are correlated, permuting one still leaves the other intact, so
  the model can partially recover. This leads to each appearing less important
  than they collectively are.
- Importance reflects *model reliance*, not causal effect. A feature the model
  uses heavily might have that relationship for reasons specific to this dataset
  or this architecture.
- Scores are post-hoc and specific to this fitted model. A different model with
  different hyperparameters could produce a somewhat different importance
  ranking.

---

## Keras vs. PyTorch: A Summary

Here is a consolidated comparison of the features we used today, with their
PyTorch equivalents:

| Feature | Keras | PyTorch |
|---|---|---|
| Early stopping | `callback_early_stopping()` in `set_engine()` | Write your own loop logic or use Lightning |
| Validation split | `validation_split = 0.2` argument | Split `DataLoader` manually |
| Learning rate scheduling | `callback_reduce_lr_on_plateau()` | `torch.optim.lr_scheduler.*` |
| Training history | Returned automatically from `fit()` | Append to list in training loop |
| TensorBoard logging | `callback_tensorboard()` | `torch.utils.tensorboard.SummaryWriter` |
| Model checkpointing | `callback_model_checkpoint()` | Manual `torch.save()` in loop |
| Softmax + cross-entropy output | Automatic for multiclass | Specify `nn.Softmax` + `nn.CrossEntropyLoss` explicitly |

The pattern is consistent: Keras externalizes infrastructure that PyTorch
internalizes. Keras is narrower but faster to stand up; PyTorch is more
expressive but requires more code. For applied research where model architecture
is standard and the domain problem is the hard part, Keras's defaults often get
you where you need to go.

---

## Next Steps

In the next lecture we will take the fitted model and use it for **subgroup
analysis** by race and gender — the variables we deliberately withheld from
the predictor set. We will look at whether the model's errors are distributed
evenly across groups or concentrated in particular subpopulations, and discuss
what that means for questions of algorithmic fairness in educational contexts.

---
