---
title: "Lecture 19: Subgroup Analysis and Algorithmic Fairness"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.width = 7,
  fig.height = 5
)

# Set to FALSE after first run to skip expensive computations
fit_model <- FALSE
```

## Overview

In Lecture 18 we fit a multinomial neural network to predict college major
field of study from ELS:2002, and we made one deliberate design choice: **race
and gender were withheld as predictors**. We assigned them the role `"subgroup"`
in the recipe and left them entirely out of the model matrix.

Today we use them — not as predictors, but as lenses for evaluating what the
model learned. This is called **subgroup analysis**, and it is the standard
first step in auditing a predictive model for disparate impact. The core
question is not whether the model is accurate on average, but whether its
errors are evenly distributed across groups or concentrated in particular
populations.

By the end of this lecture you will be able to:

1. Recover subgroup identifiers from the test set and join them to model
   predictions.
2. Decompose overall performance metrics by race, gender, and the intersection
   of the two.
3. Visualize and interpret disparities in accuracy, ROC AUC, precision, recall,
   and calibration.
4. Distinguish between different formal definitions of fairness and explain why
   they cannot all hold simultaneously.


---

## Setup and Libraries

```{r libraries}
library(tidyverse)
library(haven)
library(here)
library(janitor)
library(tidymodels)
library(kableExtra)
library(patchwork)  # composing multi-panel plots
```

```{r theme}
theme_els <- function() {
  theme_minimal(base_size = 11) +
    theme(
      plot.title    = element_text(face = "bold", size = 12),
      plot.subtitle = element_text(color = "grey40", size = 10),
      axis.title    = element_text(size = 10),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
}
```

---

## Recap: Why We Withheld Race and Gender

Recall from Lecture 18:

- **Data leakage prevention** was the primary motivation for excluding declared
  major. Race and gender are different — there is no temporal leakage issue.
- **Design choice**: we chose not to use them as predictors because we want
  the model's predictions to be based on academic preparation, attitudes, and
  postsecondary behavior — not identity characteristics.
- **Subgroup analysis role**: by withholding them from the predictor set, we
  can ask a clean counterfactual question: *does a model that never saw race
  or gender still produce predictions that vary systematically by those
  characteristics?*

If the answer is yes — and it typically is — that tells us something important
about what the model actually learned, and about the structure of the data it
was trained on.

This connects directly to the case studies you worked through in Lecture 13.
In Case Study 1 (enrollment prediction), you had to decide whether to include
student demographics as predictors. In Case Study 4 (dropout early warning),
you had to decide how to prevent disproportionate flagging of certain groups.
In each of those discussions, the question was not simply technical — it was
normative. Today we formalize that.

---

## Loading Artifacts from Lecture 18

```{r load-artifacts}
analytic_sample <- readRDS(here::here("analytic_sample.rds"))
final_fit       <- readRDS(here::here("final_fit.rds"))
cip_labels      <- read_csv(here::here("cip2_labels.csv"),
                             show_col_types = FALSE)

label_lookup <- cip_labels %>%
  select(cip2_collapsed_code, cip2_collapsed_label) %>%
  distinct() %>%
  mutate(cip2_collapsed_code = as.factor(cip2_collapsed_code))
```

---

## ELS:2002 Coding for Subgroup Variables

Before we analyze subgroup performance, we need to understand the ELS coding
schemes.

### Race (`byrace`)

ELS:2002 uses a composite race/ethnicity variable. The public-use file codes it
as follows (see NCES codebook, variable `byrace`):

| Code | Label |
|------|-------|
| 1 | American Indian / Alaska Native, non-Hispanic |
| 2 | Asian, Hawaii/Pacific Islander, non-Hispanic |
| 3 | Black or African-American, non-Hispanic |
| 4 | Hispanic, no race specified |
| 5 | Hispanic, race specified |
| 6 | Multiracial, non-Hispanic |
| 7 | White, non-Hispanic |

For analytic clarity we collapse codes 4 and 5 into a single Hispanic category
and relabel for readability. Cells with very small $n$ will be noted in the
output.

### Sex (`bysex`)

| Code | Label |
|------|-------|
| 1 | Male |
| 2 | Female |

```{r recode-subgroups}
recode_subgroups <- function(df) {
  df %>%
    mutate(
      race_label = case_when(
        byrace == 1 ~ "Am. Indian/AK Native",
        byrace == 2 ~ "Asian/Pacific Islander",
        byrace == 3 ~ "Black",
        byrace %in% c(4, 5) ~ "Hispanic",
        byrace == 6 ~ "Multiracial",
        byrace == 7 ~ "White",
        TRUE        ~ NA_character_
      ),
      sex_label = case_when(
        bysex == 1 ~ "Male",
        bysex == 2 ~ "Female",
        TRUE       ~ NA_character_
      ),
      race_label = factor(race_label, levels = c(
        "Am. Indian/AK Native", "Asian/Pacific Islander",
        "Black", "Hispanic", "Multiracial", "White"
      )),
      sex_label = factor(sex_label, levels = c("Male", "Female"))
    )
}
```

---

## Assembling the Subgroup-Annotated Test Set

The test predictions from `last_fit()` do not include `bysex` or `byrace`
because we assigned them the role `"subgroup"` in the recipe — they were never
baked into the model matrix. We recover them by joining back on `stu_id`.

```{r assemble-test-preds}
# Recover test-set rows from the analytic sample
set.seed(2002)
els_split <- initial_split(analytic_sample, prop = 0.80, strata = cip2)
els_test  <- testing(els_split)

# Subgroup metadata for the test set
subgroup_meta <- els_test %>%
  select(stu_id, bysex, byrace, cip2_collapsed_label) %>%
  recode_subgroups()

# Join to model predictions
test_preds <- collect_predictions(final_fit) %>%
  bind_cols(els_test %>% select(stu_id)) %>%
  left_join(subgroup_meta, by = "stu_id")

glimpse(test_preds)
```

### Face Validity Check

```{r face-validity-subgroup}
# Are the subgroup distributions in the test set reasonable?
test_preds %>%
  count(race_label) %>%
  mutate(pct = n / sum(n)) %>%
  kable(digits = 3, caption = "Race distribution in test set") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))

test_preds %>%
  count(sex_label) %>%
  mutate(pct = n / sum(n)) %>%
  kable(digits = 3, caption = "Sex distribution in test set") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

Small cells (particularly American Indian/Alaska Native) should be interpreted
with caution. We will note cell sizes throughout.

---

## Subgroup Analysis by Sex

### Overall Accuracy by Sex

```{r accuracy-by-sex}
accuracy_by_sex <- test_preds %>%
  filter(!is.na(sex_label)) %>%
  group_by(sex_label) %>%
  accuracy(truth = cip2, estimate = .pred_class) %>%
  select(sex_label, .estimate) %>%
  rename(accuracy = .estimate)

accuracy_by_sex %>%
  kable(digits = 3, caption = "Overall accuracy by sex") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

Accuracy alone is insufficient — we need to understand *which* majors are
being correctly classified for each group. A model could have equal overall
accuracy for men and women but still systematically misclassify one group
into different major categories than the other.

### Confusion Matrices by Sex

```{r conf-mat-by-sex, fig.width=14, fig.height=7}
# Compute confusion matrix separately for each sex
make_conf_mat_plot <- function(df, group_label) {
  conf_mat(df, truth = cip2, estimate = .pred_class) %>%
    autoplot(type = "heatmap") +
    scale_fill_gradient(low = "white", high = "#2C7BB6") +
    labs(
      title    = group_label,
      subtitle = paste0("n = ", nrow(df))
    ) +
    theme_els() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
      axis.text.y = element_text(size = 7),
      legend.position = "none"
    )
}

conf_plots_sex <- test_preds %>%
  filter(!is.na(sex_label)) %>%
  group_by(sex_label) %>%
  group_map(~ make_conf_mat_plot(.x, .y$sex_label), .keep = TRUE)

conf_plots_sex[[1]] + conf_plots_sex[[2]] +
  plot_annotation(
    title    = "Confusion Matrices by Sex",
    subtitle = "Rows: true class  |  Columns: predicted class",
    theme    = theme_els()
  )
```

### Per-Class F1 by Sex

```{r f1-by-sex}
f1_by_sex <- test_preds %>%
  filter(!is.na(sex_label)) %>%
  group_by(sex_label) %>%
  f_meas(truth = cip2, estimate = .pred_class, estimator = "macro") %>%
  select(sex_label, .estimate) %>%
  rename(f1 = .estimate)

f1_by_sex %>%
  kable(digits = 3, caption = "Macro-averaged F1 by sex") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

```{r f1-by-sex-major, fig.height=7}
# F1 broken out by major field and sex — shows where gaps arise
f1_by_sex_major <- test_preds %>%
  filter(!is.na(sex_label)) %>%
  group_by(sex_label) %>%
  group_modify(function(df, grp) {
    df %>%
      group_by(cip2) %>%
      summarise(
        n         = n(),
        precision = precision_vec(truth = cip2, estimate = .pred_class),
        recall    = recall_vec(truth = cip2, estimate = .pred_class),
        f1        = f_meas_vec(truth = cip2, estimate = .pred_class),
        .groups   = "drop"
      )
  }) %>%
  left_join(label_lookup, by = join_by(cip2 == cip2_collapsed_code))

f1_by_sex_major %>%
  filter(!is.na(cip2_collapsed_label)) %>%
  mutate(cip2_collapsed_label = fct_reorder(cip2_collapsed_label, f1)) %>%
  ggplot(aes(x = f1, y = cip2_collapsed_label, fill = sex_label)) +
  geom_col(position = "dodge", alpha = 0.85) +
  scale_fill_manual(values = c(Male = "#2C7BB6", Female = "#D7191C")) +
  scale_x_continuous(limits = c(0, 1)) +
  labs(
    title    = "Per-Major F1 Score by Sex",
    subtitle = "Higher is better; gaps indicate disparate classification accuracy by major",
    x        = "F1 Score",
    y        = NULL,
    fill     = "Sex"
  ) +
  theme_els()
```

Pay attention to majors where there is a large gap between bars. If the model
consistently underperforms for women in STEM fields or for men in Education,
that pattern is meaningful — and it almost certainly reflects both real
differences in the training data distribution and the model's tendency to
amplify those patterns.

### ROC AUC by Sex

```{r roc-auc-by-sex}
roc_auc_by_sex <- test_preds %>%
  filter(!is.na(sex_label)) %>%
  group_by(sex_label) %>%
  roc_auc(
    truth      = cip2,
    starts_with(".pred_"),
    estimator  = "macro_weighted"
  ) %>%
  select(sex_label, .estimate) %>%
  rename(roc_auc = .estimate)

roc_auc_by_sex %>%
  kable(digits = 3, caption = "Macro-weighted ROC AUC by sex") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

---

## Subgroup Analysis by Race

### Overall Accuracy by Race

```{r accuracy-by-race}
accuracy_by_race <- test_preds %>%
  filter(!is.na(race_label)) %>%
  group_by(race_label) %>%
  summarise(
    n        = n(),
    accuracy = accuracy_vec(truth = cip2, estimate = .pred_class),
    .groups  = "drop"
  ) %>%
  arrange(desc(accuracy))

accuracy_by_race %>%
  kable(digits = 3, caption = "Overall accuracy by race (test set)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

```{r accuracy-race-plot, fig.height=5}
accuracy_by_race %>%
  filter(n >= 30) %>%        # suppress small-cell estimates
  mutate(race_label = fct_reorder(race_label, accuracy)) %>%
  ggplot(aes(x = accuracy, y = race_label)) +
  geom_col(fill = "#2C7BB6", alpha = 0.85) +
  geom_text(aes(label = paste0("n=", n)), hjust = -0.15, size = 3.2) +
  scale_x_continuous(limits = c(0, 0.6), labels = scales::percent) +
  labs(
    title    = "Classification Accuracy by Race",
    subtitle = "ELS:2002 test set  |  Groups with n < 30 suppressed",
    x        = "Accuracy",
    y        = NULL
  ) +
  theme_els()
```

### ROC AUC by Race

```{r roc-auc-by-race}
roc_auc_by_race <- test_preds %>%
  filter(!is.na(race_label)) %>%
  group_by(race_label) %>%
  group_modify(function(df, grp) {
    if (nrow(df) < 30) return(tibble(.estimate = NA_real_))
    df %>%
      roc_auc(
        truth     = cip2,
        starts_with(".pred_"),
        estimator = "macro_weighted"
      ) %>%
      select(.estimate)
  }) %>%
  rename(roc_auc = .estimate) %>%
  left_join(accuracy_by_race %>% select(race_label, n), by = "race_label") %>%
  filter(!is.na(roc_auc))

roc_auc_by_race %>%
  arrange(desc(roc_auc)) %>%
  kable(digits = 3, caption = "Macro-weighted ROC AUC by race (groups with n >= 30)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

```{r roc-auc-race-plot, fig.height=5}
roc_auc_by_race %>%
  mutate(race_label = fct_reorder(race_label, roc_auc)) %>%
  ggplot(aes(x = roc_auc, y = race_label)) +
  geom_col(fill = "#2C7BB6", alpha = 0.85) +
  geom_text(aes(label = paste0("n=", n)), hjust = -0.15, size = 3.2) +
  scale_x_continuous(limits = c(0, 1),
                     labels = scales::number_format(accuracy = 0.01)) +
  labs(
    title    = "Macro-Weighted ROC AUC by Race",
    subtitle = "ELS:2002 test set  |  Groups with n < 30 suppressed",
    x        = "ROC AUC",
    y        = NULL
  ) +
  theme_els()
```

### Predicted Major Distribution by Race

Even if accuracy were identical across groups, the *distribution of
predictions* might differ systematically. If the model consistently routes
Black and Hispanic students toward certain major categories more than White
or Asian students with similar observed characteristics, that is a form of
disparate impact regardless of whether those predictions are correct.

```{r pred-dist-by-race, fig.height=8}
test_preds %>%
  filter(!is.na(race_label)) %>%
  left_join(
    label_lookup %>% rename(pred_major_label = cip2_collapsed_label),
    by = join_by(.pred_class == cip2_collapsed_code)
  ) %>%
  count(race_label, pred_major_label) %>%
  group_by(race_label) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup() %>%
  filter(!is.na(pred_major_label)) %>%
  ggplot(aes(x = pct, y = fct_reorder(pred_major_label, pct), fill = race_label)) +
  geom_col(position = "dodge", alpha = 0.85) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title    = "Distribution of Predicted Majors by Race",
    subtitle = "Each row is a predicted major; bars show % of each racial group predicted there",
    x        = "% of group",
    y        = NULL,
    fill     = "Race"
  ) +
  theme_els() +
  theme(legend.position = "right")
```

---

## Intersectional Analysis: Race × Gender

Subgroup analyses by race alone and by gender alone can mask important
variation. A model that performs well for White women and Black men might still
perform poorly for Black women — a pattern invisible in the marginal analyses.
This is the core insight of **intersectionality** in a fairness context.

```{r intersectional-accuracy, fig.height=6}
intersect_accuracy <- test_preds %>%
  filter(!is.na(race_label), !is.na(sex_label)) %>%
  group_by(race_label, sex_label) %>%
  summarise(
    n        = n(),
    accuracy = accuracy_vec(truth = cip2, estimate = .pred_class),
    .groups  = "drop"
  ) %>%
  filter(n >= 20)   # suppress very small cells

intersect_accuracy %>%
  ggplot(aes(x = race_label, y = accuracy, fill = sex_label)) +
  geom_col(position = "dodge", alpha = 0.85) +
  geom_text(
    aes(label = paste0("n=", n)),
    position = position_dodge(width = 0.9),
    vjust = -0.4, size = 2.8
  ) +
  scale_fill_manual(values = c(Male = "#2C7BB6", Female = "#D7191C")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 0.7)) +
  labs(
    title    = "Classification Accuracy: Race × Sex Interaction",
    subtitle = "ELS:2002 test set  |  Cells with n < 20 suppressed",
    x        = NULL,
    y        = "Accuracy",
    fill     = "Sex"
  ) +
  theme_els() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r intersectional-roc, fig.height=6}
intersect_roc <- test_preds %>%
  filter(!is.na(race_label), !is.na(sex_label)) %>%
  group_by(race_label, sex_label) %>%
  group_modify(function(df, grp) {
    if (nrow(df) < 30) return(tibble(roc_auc = NA_real_, n = nrow(df)))
    roc <- df %>%
      roc_auc(
        truth     = cip2,
        starts_with(".pred_"),
        estimator = "macro_weighted"
      )
    tibble(roc_auc = roc$.estimate, n = nrow(df))
  }) %>%
  filter(!is.na(roc_auc))

intersect_roc %>%
  ggplot(aes(x = race_label, y = roc_auc, fill = sex_label)) +
  geom_col(position = "dodge", alpha = 0.85) +
  geom_text(
    aes(label = paste0("n=", n)),
    position = position_dodge(width = 0.9),
    vjust = -0.4, size = 2.8
  ) +
  scale_fill_manual(values = c(Male = "#2C7BB6", Female = "#D7191C")) +
  scale_y_continuous(limits = c(0, 1),
                     labels = scales::number_format(accuracy = 0.01)) +
  labs(
    title    = "ROC AUC: Race × Sex Interaction",
    subtitle = "Macro-weighted  |  Cells with n < 30 suppressed",
    x        = NULL,
    y        = "ROC AUC",
    fill     = "Sex"
  ) +
  theme_els() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

---

## What the Model Actually Learned: Predicted Probabilities by Race

Looking at hard classifications (`.pred_class`) tells us about errors. But the
*predicted probabilities* reveal something additional: how *confident* the model
is, and whether that confidence is calibrated differently across groups.

```{r max-prob-by-race, fig.height=5}
# Max predicted probability = how confident is the model in its top choice?
test_preds %>%
  filter(!is.na(race_label)) %>%
  mutate(
    max_prob = pmap_dbl(
      select(., starts_with(".pred_")),
      ~ max(c(...))
    ),
    correct = (cip2 == .pred_class)
  ) %>%
  ggplot(aes(x = max_prob, fill = race_label)) +
  geom_density(alpha = 0.45) +
  scale_fill_brewer(palette = "Set2") +
  facet_wrap(~ correct, labeller = as_labeller(c(`TRUE` = "Correct", `FALSE` = "Incorrect"))) +
  labs(
    title    = "Model Confidence (Max Predicted Probability) by Race",
    subtitle = "Separated by whether the prediction was correct",
    x        = "Max predicted probability",
    y        = "Density",
    fill     = "Race"
  ) +
  theme_els()
```

If some racial groups have systematically lower max probabilities even on correct
predictions, the model is less certain about those students. That uncertainty
has practical consequences: a downstream decision-maker using probability
thresholds would treat those students differently.

---

## Connecting to Major-Specific Disparities

We have seen overall performance gaps. Now we zoom in: for which specific major
categories does the model fail most for underrepresented groups? This is where
subgroup analysis connects directly to the substantive question — who gets
routed where.

```{r recall-by-race-major, fig.height=9}
# Per-class recall by race
recall_by_race_major <- test_preds %>%
  filter(!is.na(race_label)) %>%
  group_by(race_label) %>%
  group_modify(function(df, grp) {
    if (nrow(df) < 30) return(tibble())
    df %>%
      group_by(cip2) %>%
      summarise(
        n_true    = n(),
        n_correct = sum(.pred_class == cip2),
        recall    = n_correct / n_true,
        .groups   = "drop"
      )
  }) %>%
  left_join(label_lookup, by = join_by(cip2 == cip2_collapsed_code)) %>%
  filter(!is.na(cip2_collapsed_label))

recall_by_race_major %>%
  filter(n_true >= 10) %>%   # suppress very small within-group cells
  mutate(cip2_collapsed_label = fct_reorder(cip2_collapsed_label, recall)) %>%
  ggplot(aes(x = recall, y = cip2_collapsed_label, fill = race_label)) +
  geom_col(position = "dodge", alpha = 0.85) +
  scale_fill_brewer(palette = "Set2") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1)) +
  labs(
    title    = "Per-Major Recall by Race",
    subtitle = "Among true members of each major, what fraction did the model correctly identify?",
    x        = "Recall",
    y        = NULL,
    fill     = "Race"
  ) +
  theme_els() +
  theme(legend.position = "right")
```

---

## Formal Fairness Criteria: An Impossible Dream?

The empirical patterns above motivate a deeper question: what does it mean for
a model to be *fair*? Researchers in algorithmic fairness have proposed several
formal criteria, and an important mathematical result shows that most of them
cannot hold simultaneously when base rates differ across groups.

### Three common criteria

**Demographic parity** (independence): The probability of receiving a predicted
classification is equal across groups. If the model predicts "STEM" for 20% of
White students, it should predict "STEM" for 20% of Black students too.

**Equalized odds**: The true positive rate (recall) and false positive rate are
equal across groups. The model is equally good at detecting true STEM majors,
and equally cautious about false alarms, regardless of race.

**Calibration** (sufficiency): Among students to whom the model assigns a 70%
probability of being a STEM major, 70% should actually be STEM majors —
regardless of race. The model's confidence is accurate for all groups.

### Why they conflict

When the outcome base rates differ across groups — and they do in ELS:2002 —
you provably cannot satisfy demographic parity, equalized odds, and perfect
calibration at the same time (Chouldechova, 2017; Kleinberg et al., 2017). This
is not a limitation of our model or our implementation. It is a mathematical
theorem about the structure of the problem.

The practical implication: choosing a fairness criterion is a normative
decision, not a technical one. A researcher building a major prediction tool
must decide which kind of error — false inclusions or false exclusions from
certain major pathways — is more costly, and for whom. That decision cannot be
made by optimizing a loss function. It requires normative reasoning about the
purposes of prediction and the context in which it will be used.

```{r fairness-criteria-check}
# Do base rates differ across racial groups?
# (This motivates why the impossibility result applies here)
analytic_sample %>%
  recode_subgroups() %>%
  filter(!is.na(race_label)) %>%
  count(race_label, cip2) %>%
  group_by(race_label) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup() %>%
  left_join(label_lookup, by = join_by(cip2 == cip2_collapsed_code)) %>%
  filter(!is.na(cip2_collapsed_label)) %>%
  select(race_label, cip2_collapsed_label, pct) %>%
  pivot_wider(names_from = race_label, values_from = pct, values_fill = 0) %>%
  kable(digits = 3,
        caption = "Major field base rates by race — these differ, so fairness criteria conflict") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"))
```

---

## What Is the Source of the Disparity?

When a model produces unequal performance across groups, there are several
possible explanations. These are not mutually exclusive.

**Representation in training data.** If Black or Hispanic students are
underrepresented in the training data, the model has seen fewer examples of
their patterns and will perform worse on them at test time. This is a
*data problem*, not a model problem — but the model amplifies it.

**Proxy features.** We excluded race and gender from the predictor set. But
many of the predictors we did include — parental education, income, school
sector, regional location — are correlated with race. The model can
reconstruct something like race from these proxies. This is called proxy
discrimination, and it is one of the core challenges of fairness-aware
machine learning.

**Label quality.** The outcome — degree major as recorded in postsecondary
transcripts — may be measured differently or with different error rates across
groups. If certain students are more likely to switch majors after initial
enrollment, their transcript major may be less stable, adding noise to the
labels for that group.

**Structural inequality in the predictors themselves.** A Black student and a
White student with the same math test score are not in the same position if
they attended schools with very different resource levels, course offerings, or
grading standards. The test score means something different in context, but the
model treats it as a number. This is the deepest problem — it applies even if
we solve the data and proxy issues.

```{r proxy-check, fig.height=5}
# Do the proxy variables vary by race? This is expected and not surprising,
# but it motivates why excluding race doesn't "solve" the problem.
analytic_sample %>%
  recode_subgroups() %>%
  filter(!is.na(race_label)) %>%
  group_by(race_label) %>%
  summarise(
    mean_ses  = mean(byses2, na.rm = TRUE),
    mean_math = mean(bytxmstd, na.rm = TRUE),
    mean_read = mean(bytxrstd, na.rm = TRUE),
    n         = n(),
    .groups   = "drop"
  ) %>%
  filter(n >= 30) %>%
  pivot_longer(c(mean_ses, mean_math, mean_read),
               names_to = "measure", values_to = "mean_value") %>%
  mutate(
    measure = recode(measure,
                     mean_ses  = "SES composite (byses2)",
                     mean_math = "Math test score (bytxmstd)",
                     mean_read = "Reading test score (bytxrstd)")
  ) %>%
  ggplot(aes(x = race_label, y = mean_value, fill = race_label)) +
  geom_col(alpha = 0.85, show.legend = FALSE) +
  facet_wrap(~ measure, scales = "free_y") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title    = "Mean Values of Key Proxy Predictors by Race",
    subtitle = "These predictors are in the model; their variation by race enters model predictions",
    x        = NULL,
    y        = "Mean"
  ) +
  theme_els() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1, size = 8))
```

This plot is important. It shows that the variables we kept in the model vary
substantially by race. The model did not "see" race directly — but it saw the
world through variables that are partially constituted by race-based inequality.
That information does not disappear because we excluded `byrace` from the
recipe.

---

## Connecting to Lecture 13: The Case Studies Revisited

In Lecture 13 you worked through several scenarios involving predictive models
in educational contexts. The subgroup analysis above is the empirical version of
the normative questions you were asked to consider.

**Case Study 1 (Enrollment prediction):** Your group had to decide whether to
include student demographics as predictors. We showed above that excluding them
does not eliminate their influence — it just makes that influence opaque. An
enrollment model that systematically overestimates yield for high-income students
and underestimates it for low-income students will lead to aid misallocation
regardless of whether income was in the model explicitly.

**Case Study 2 (School enrollment forecasting):** The question of which schools
are predicted to decline often correlates heavily with the racial composition
of the school's attendance zone. A model that uses demographic trends as
predictors — which is technically useful — can also encode and amplify
residential segregation patterns into resource allocation decisions.

**Case Study 4 (Dropout early warning):** This is the sharpest case. An early
warning system that flags Black and Hispanic students at higher rates than White
students with similar observable characteristics — even without explicit race
coding — can create a self-fulfilling prophecy: students who are flagged receive
intrusive interventions that may themselves affect outcomes, and students who are
not flagged are left without support. The system's errors are not random noise;
they are structured by inequality.

The through-line: what you include in a model, what you exclude, and for what
purpose you deploy the model are not value-neutral technical choices. The
subgroup analysis gives you the empirical tools to surface these issues. What
you do with those results requires judgment that the numbers cannot supply.

---

## Reporting Subgroup Performance: A Practical Template

When publishing or deploying a model with real-world consequences, subgroup
performance reporting is increasingly expected — and in some regulatory
contexts, required. A minimal responsible reporting template includes:

1. **Subgroup N's and proportions** in train and test sets.
2. **Accuracy, precision, recall, and F1** disaggregated by relevant subgroups.
3. **ROC AUC** disaggregated, with note on small-cell suppression thresholds.
4. **Calibration plots** by subgroup (not covered today but important for
   high-stakes deployment).
5. **Intersectional breakdowns** for the most policy-relevant combinations.
6. **Qualitative narrative** about the sources of any gaps observed and what
   the model cannot adjust for (e.g., structural inequality in predictors).

```{r summary-table}
# Compact summary table for reporting
summary_race <- test_preds %>%
  filter(!is.na(race_label)) %>%
  group_by(group = as.character(race_label)) %>%
  summarise(
    subgroup_type = "Race",
    n        = n(),
    accuracy = accuracy_vec(truth = cip2, estimate = .pred_class),
    .groups  = "drop"
  )

summary_sex <- test_preds %>%
  filter(!is.na(sex_label)) %>%
  group_by(group = as.character(sex_label)) %>%
  summarise(
    subgroup_type = "Sex",
    n        = n(),
    accuracy = accuracy_vec(truth = cip2, estimate = .pred_class),
    .groups  = "drop"
  )

summary_table <- bind_rows(summary_race, summary_sex) %>%
  filter(n >= 20) %>%
  arrange(subgroup_type, desc(accuracy))

summary_table %>%
  select(subgroup_type, group, n, accuracy) %>%
  kable(digits = 3,
        caption = "Subgroup performance summary (accuracy; groups with n < 20 suppressed)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed")) %>%
  pack_rows(index = table(summary_table$subgroup_type))
```
