---
title: "Conformal Prediction for Individual-Level Uncertainty"
author: "Will Doyle"
date: "2025-04-01"
output: github_document
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(janitor)
library(probably)   # conformal inference via tidymodels
library(xgboost)
```

---

# Part 1: The Problem with Point Predictions

Peviously we fit an XGBoost model to predict whether a student passes an online course in the Open University Learning Analytics Dataset (OULAD). Our best model produced a predicted probability for every student — something like 0.73 for one student, 0.41 for another.

But that number is incomplete. It answers "what does the model think?" without answering "how much should we trust that answer?" A 0.73 from a well-calibrated model on a familiar type of student is a very different thing from a 0.73 on an unusual student where the model is extrapolating.

Standard practice is to use the model's internal class probabilities directly as uncertainty estimates. For complex, non-parametric models like XGBoost — which make no assumptions about the data-generating process — this is problematic. The model's internal probabilities may be **miscalibrated**: systematically too confident in some regions of the feature space and too uncertain in others. There is no easy way to audit them from the model alone.

**Conformal prediction** offers a principled alternative. Rather than trusting the model's internal uncertainty, it uses empirical evidence from held-out data to construct prediction sets with a rigorous, distribution-free guarantee: the true outcome is contained in the prediction set at least $(1 - \alpha)$% of the time — for any underlying model, regardless of its parametric form, and without requiring the data to follow any particular distribution.

For individual students facing high-stakes predictions (intervention decisions, early warnings, resource allocation), that distinction matters.

---

# Part 2: Conformal Prediction Theory

## The Nonconformity Score

Every conformal procedure is built around a nonconformity score — a number that measures how surprising a new observation is relative to the training data, as judged by the model.

For a binary classification problem like ours, the natural score is:

$$s_i = 1 - \hat{p}(y_i)$$

where $\hat{p}(y_i)$ is the predicted probability the model assigns to the true class label. If the model correctly and confidently predicts student $i$ passed, then $\hat{p}(\text{passed})$ is close to 1 and $s_i \approx 0$. If the model was wrong or uncertain, $s_i$ is large.

A large nonconformity score says: this observation does not conform well to what the model has learned — it is "surprising" given the training evidence.

## The Calibration Set and Quantile Threshold

The key to conformal inference is evaluating these scores on data the model has never seen — a held-out calibration set. The procedure:

1. Fit the model on a training set.
2. Compute $s_1, \ldots, s_n$ on the calibration set.
3. For target coverage $1 - \alpha$, find the threshold:

$$\hat{q} = \text{Quantile}\!\left(\{s_1, \ldots, s_n\},\ \frac{\lceil (n+1)(1-\alpha) \rceil}{n}\right)$$

4. For a new test observation, include class $k$ in the prediction set if:

$$1 - \hat{p}(k) \leq \hat{q}$$

That is, include any class whose nonconformity score falls at or below the threshold.

The $\frac{n+1}{n}$ correction accounts for the fact that we need coverage to hold for the new point, not just the calibration points. As $n \to \infty$ it vanishes.

## Why the Guarantee Holds

The guarantee rests on the assumption of exchangeability — roughly, that the calibration observations and the new test observation are drawn from the same distribution. Under this condition, the new observation's score is equally likely to land anywhere in the distribution of calibration scores, so the probability it exceeds the $(1-\alpha)$th quantile is at most $\alpha$. 

This is a finite-sample, distribution-free result. No normality. No parametric assumptions. No requirement that the model be well-specified. The model is just a black box that produces a nonconformity score.

## What the Output Looks Like

For a binary outcome, each test student receives one of three prediction set types:

| Set | Meaning |
|---|---|
| `{passed}` | Model is confident this student will pass; only "passed" is plausible |
| `{not_passed}` | Model is confident this student will not pass |
| `{passed, not_passed}` | Model is uncertain; both outcomes are plausible given the evidence |
| `{}` | Rare anomaly — model assigns low probability to all classes; flags unusual observation |

The size of the prediction set is a direct, honest measure of individual-level uncertainty: a singleton set is a confident prediction; a both-class set is an honest "I don't know."

---

# Part 3: Data Setup

We pick up from where the XGBoost lecture left off — the OULAD dataset, binary outcome of whether a student passed their online course.

```{r load_data}
ou <- read_csv("oulad.csv") %>%
  mutate(result = fct_relevel(as_factor(result), c("passed", "not_passed"))) %>%
  select(-final_result)
```

For conformal prediction we need three distinct data partitions:

- Training set: fit the XGBoost model
- Calibration set: compute nonconformity scores, determine $\hat{q}$
- Test set: evaluate coverage and inspect individual predictions

```{r split_data}
# 60% training + calibration, 40% test
ou_split      <- initial_split(ou, prop = 0.6)
ou_train_cal  <- training(ou_split)
ou_test       <- testing(ou_split)

# Within the training portion: 75% train, 25% calibration
cal_split <- initial_split(ou_train_cal, prop = 0.75)
ou_train  <- training(cal_split)
ou_cal    <- testing(cal_split)
```

## Recipe and Model Specification

Same recipe and model spec as the XGBoost lecture:

```{r}
xgb_grid <- grid_space_filling(
  tree_depth(),
  min_n(),
  loss_reduction(),
  sample_size = sample_prop(),
  finalize(mtry(), ou_train),
  learn_rate(),
  size = 30
)
```


```{r}
xgb_spec <- boost_tree(
  trees        = 100,
  tree_depth   = tune(),   # controls tree complexity
  min_n        = tune(),   # minimum observations per node
  loss_reduction = tune(), # γ: minimum gain to accept a split
  sample_size  = tune(),   # row subsampling fraction
  mtry         = tune(),   # column subsampling per split
  learn_rate   = tune()    # η: step size
) %>%
  set_engine("xgboost") %>%
  set_mode("classification")
```


```{r recipe}
ou_rec <- recipe(result ~ ., ou_train) %>%
  update_role(id_student, new_role = "id") %>%
  step_other(all_nominal_predictors(), threshold = 0.05) %>%
  step_unknown(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_predictors())
```

```{r}
ou_wf <- workflow() %>%
  add_recipe(ou_rec) %>%
  add_model(xgb_spec)
```



```{r fit_model}
fit_model <- FALSE

if (fit_model) {
  xg_tune_res <- tune_grid(
    ou_wf,
    grid      = xgb_grid,
    resamples = ou_rs
  )
  save(xg_tune_res, file = "xg_tune_res.Rdata")
} else {
  load("xg_tune_res.Rdata")
}
```

```{r}

best_auc <- select_best(xg_tune_res, metric = "roc_auc")

final_xgb <- finalize_workflow(ou_wf, best_auc)

xgb_fit<-fit(final_xgb,ou_train)

```

---

# Part 4: Implementing Conformal Prediction from Scratch

Working through the mechanics by hand makes the procedure transparent before we let a package automate it.

## Step 1: Predict on the Calibration Set

```{r cal_preds}
cal_preds <- augment(xgb_fit, ou_cal)

cal_preds %>%
  select(result, .pred_passed, .pred_not_passed) %>%
  head(8)
```

## Step 2: Compute Nonconformity Scores

For each calibration student, the score is $1 - \hat{p}(\text{true class})$:

```{r cal_scores}
cal_scores <- cal_preds %>%
  mutate(
    true_class_prob = if_else(result == "passed", .pred_passed, .pred_not_passed),
    score           = 1 - true_class_prob
  )

cal_scores %>%
  select(result, .pred_passed, .pred_not_passed, true_class_prob, score) %>%
  head(10)
```

## Step 3: Find the Quantile Threshold

```{r quantile_threshold}
alpha   <- 0.10   # target: 90% coverage
n       <- nrow(cal_scores)
q_level <- min(ceiling((n + 1) * (1 - alpha)) / n, 1)
q_hat   <- quantile(cal_scores$score, probs = q_level)

cat("Calibration set size:", n, "\n")
cat("Quantile level used: ", round(q_level, 4), "\n")
cat("Score threshold q̂:  ", round(q_hat, 4), "\n")
```

Let's visualize the distribution of calibration scores and where $\hat{q}$ falls:

```{r score_distribution}
ggplot(cal_scores, aes(x = score)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white", alpha = 0.85) +
  geom_vline(xintercept = q_hat, color = "firebrick", linewidth = 1.2, linetype = "dashed") +
  annotate(
    "text", x = q_hat + 0.02, y = Inf, vjust = 2,
    label = paste0("q\u0302 = ", round(q_hat, 3)),
    color = "firebrick", size = 4
  ) +
  facet_wrap(~result, scales = "free_y") +
  labs(
    title    = "Distribution of calibration nonconformity scores by true outcome",
    subtitle = paste0("Scores above the dashed line exceed the ", (1 - alpha) * 100, "% threshold"),
    x        = "Nonconformity score  (1 \u2212 predicted probability of true class)",
    y        = "Count"
  ) +
  theme_minimal()
```

Students correctly predicted with high confidence have scores near 0. Students where the model was uncertain or wrong have scores near 1.

## Step 4: Build Prediction Sets on the Test Data

```{r test_pred_sets}
test_preds <- augment(xgb_fit, ou_test)

conformal_sets <- test_preds %>%
  mutate(
    include_passed     = (1 - .pred_passed)     <= q_hat,
    include_not_passed = (1 - .pred_not_passed) <= q_hat,
    pred_set = case_when(
      include_passed  & include_not_passed  ~ "{passed, not_passed}",
      include_passed  & !include_not_passed ~ "{passed}",
      !include_passed & include_not_passed  ~ "{not_passed}",
      TRUE                                  ~ "{}"
    )
  )

conformal_sets %>%
  count(pred_set) %>%
  mutate(pct = n / sum(n)) %>%
  arrange(desc(n))
```

## Step 5: Verify Empirical Coverage

```{r verify_coverage}
coverage_check <- conformal_sets %>%
  mutate(
    covered = case_when(
      result == "passed"     & include_passed     ~ TRUE,
      result == "not_passed" & include_not_passed ~ TRUE,
      TRUE                                        ~ FALSE
    )
  )

coverage_check %>%
  summarise(
    empirical_coverage = mean(covered),
    target_coverage    = 1 - alpha,
    n_test             = n()
  )
```

Empirical coverage should meet or exceed 90%. The theoretical guarantee is at least $(1 - \alpha)$, and deviations are sampling noise.

---

# Part 5: A Note on `probably` and Classification

The `probably` package in tidymodels provides `int_conformal_split()` and `int_conformal_cv()`, but these are designed for **regression** problems. Internally they compute a numeric interval around `.pred` — a column that doesn't exist for a classification model (which produces `.pred_passed` and `.pred_not_passed` instead). Calling `int_conformal_split()` on a classification workflow will error.

For classification conformal prediction sets, we implement the procedure ourselves, as in Part 4. This is not a limitation — it's actually the right framing. Classification conformal prediction produces sets, not numeric intervals, and the from-scratch implementation makes the logic fully transparent.

What `probably` does provide that is useful for classification is probability calibration tools: functions that correct miscalibrated class probabilities using isotonic regression or Platt scaling. These address a related but different question — not "what is the coverage guarantee?" but "are the model's raw probabilities well-calibrated?" We use conformal prediction sets as our primary uncertainty output, so we don't need this here, but it's worth knowing it exists.

```{r probably_calibration_note, eval=FALSE}
# For reference: probably's cal_estimate_logistic() can calibrate raw XGBoost
# probabilities using a logistic regression on the calibration set. This is
# a different operation from conformal prediction — it adjusts the probabilities
# themselves rather than wrapping them in a coverage-guaranteed set.

# cal_obj <- cal_estimate_logistic(cal_preds, truth = result)
# cal_apply(test_preds, cal_obj)  # returns recalibrated .pred_passed
```

---

# Part 6: Individual-Level Uncertainty

This is the core payoff. Conformal prediction doesn't just give us aggregate statistics — it tells us something meaningful about each student individually: is this prediction confident, or is the model genuinely uncertain about this person?

## Who Gets a Confident Prediction, and Who Doesn't?

```{r individual_uncertainty}
individual_uncertainty <- conformal_sets %>%
  mutate(
    set_size    = include_passed + include_not_passed,
    uncertainty = case_when(
      set_size == 2 ~ "Uncertain (both classes)",
      set_size == 0 ~ "Anomaly (empty set)",
      TRUE          ~ "Confident (singleton)"
    ),
    margin = abs(.pred_passed - .pred_not_passed)
  )

individual_uncertainty %>%
  count(uncertainty, result) %>%
  mutate(pct = n / sum(n), .by = result) %>%
  ggplot(aes(x = result, y = pct, fill = uncertainty)) +
  geom_col(position = "dodge", alpha = 0.85) +
  scale_fill_manual(
    values = c(
      "Confident (singleton)"      = "steelblue",
      "Uncertain (both classes)"   = "firebrick",
      "Anomaly (empty set)"        = "grey50"
    )
  ) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title    = "Prediction confidence by true outcome",
    subtitle = paste0("90% conformal coverage  |  q\u0302 = ", round(q_hat, 3)),
    x        = "True outcome",
    y        = "Proportion of students",
    fill     = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Confidence as a Function of Model Probability

Prediction set type is directly determined by where the model's probabilities fall relative to $\hat{q}$. Students with high predicted probabilities for one class get singleton sets; students near the 50/50 boundary get both-class sets.

```{r margin_vs_set_type}
ggplot(individual_uncertainty, aes(x = margin, fill = uncertainty)) +
  geom_histogram(bins = 40, color = "white", alpha = 0.85) +
  scale_fill_manual(
    values = c(
      "Confident (singleton)"    = "steelblue",
      "Uncertain (both classes)" = "firebrick",
      "Anomaly (empty set)"      = "grey50"
    )
  ) +
  labs(
    title    = "Prediction margin by conformal set type",
    subtitle = "Students near the decision boundary (small margin) correctly flagged as uncertain",
    x        = "| P(passed) \u2212 P(not passed) |",
    y        = "Count",
    fill     = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Student-Level Inspection: Calibrated Predictions

One of the most useful diagnostics is to sort individual test students by their predicted probability and overlay which ones get confident vs. uncertain prediction sets. This shows exactly where the model knows what it's doing and where it doesn't.

```{r student_level_plot}
individual_uncertainty %>%
  arrange(.pred_passed) %>%
  mutate(rank = row_number()) %>%
  ggplot(aes(x = rank, y = .pred_passed)) +
  geom_point(aes(color = uncertainty), alpha = 0.6, size = 0.7) +
  geom_point(
    data = ~ filter(.x, result == "passed"),
    aes(y = 1.05), shape = "|", color = "black", size = 1.5, alpha = 0.3
  ) +
  geom_hline(
    yintercept = c(1 - q_hat, q_hat),
    linetype = "dashed", color = "grey40", linewidth = 0.7
  ) +
  scale_color_manual(
    values = c(
      "Confident (singleton)"    = "steelblue",
      "Uncertain (both classes)" = "firebrick",
      "Anomaly (empty set)"      = "grey50"
    )
  ) +
  annotate("text", x = 5, y = 1 - q_hat + 0.03, label = "Upper conf. boundary",
           color = "grey40", size = 3, hjust = 0) +
  annotate("text", x = 5, y = q_hat - 0.03,     label = "Lower conf. boundary",
           color = "grey40", size = 3, hjust = 0) +
  labs(
    title    = "Predicted probability with individual-level conformal uncertainty",
    subtitle = "Students between the dashed lines receive both-class (uncertain) prediction sets | ticks = true passers",
    x        = "Students ranked by predicted P(passed)",
    y        = "Predicted P(passed)",
    color    = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The dashed lines mark the boundaries of the uncertain zone. Students between them — where the model cannot distinguish their outcome with sufficient confidence at the target coverage level — receive the honest `{passed, not_passed}` prediction set. Students outside the zone receive a confident singleton. The width of the uncertain zone is controlled by $\alpha$: tighten it (higher coverage) and the zone widens; loosen it and the zone narrows but the guarantee weakens.

---

# Part 7: The Coverage–Efficiency Tradeoff

The choice of $\alpha$ trades off two competing goods: coverage (how reliably we contain the true outcome) and efficiency (how often we give a confident, actionable singleton prediction). There is no free lunch — you cannot get both at once.

```{r coverage_tradeoff}
alpha_grid <- tibble(alpha = seq(0.05, 0.45, by = 0.05)) %>%
  mutate(
    q_val = map_dbl(alpha, function(a) {
      ql <- min(ceiling((n + 1) * (1 - a)) / n, 1)
      quantile(cal_scores$score, probs = ql)
    }),
    test_results = map(q_val, function(q) {
      test_preds %>%
        mutate(
          include_passed     = (1 - .pred_passed)     <= q,
          include_not_passed = (1 - .pred_not_passed) <= q,
          covered = case_when(
            result == "passed"     & include_passed     ~ TRUE,
            result == "not_passed" & include_not_passed ~ TRUE,
            TRUE                                        ~ FALSE
          ),
          set_size = include_passed + include_not_passed
        ) %>%
        summarise(
          empirical_coverage = mean(covered),
          pct_singleton      = mean(set_size == 1),
          pct_both           = mean(set_size == 2),
          pct_empty          = mean(set_size == 0)
        )
    })
  ) %>%
  unnest(test_results)

alpha_grid %>%
  mutate(target_coverage = 1 - alpha) %>%
  pivot_longer(
    c(empirical_coverage, pct_singleton, pct_both),
    names_to  = "metric",
    values_to = "value"
  ) %>%
  ggplot(aes(x = target_coverage, y = value, color = metric)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  scale_color_manual(
    values = c(
      "empirical_coverage" = "steelblue",
      "pct_singleton"      = "forestgreen",
      "pct_both"           = "firebrick"
    ),
    labels = c("Empirical coverage", "% singleton sets", "% both-class sets")
  ) +
  labs(
    title    = "Coverage\u2013efficiency tradeoff across \u03b1 levels",
    subtitle = "Higher required coverage forces more uncertain (both-class) prediction sets",
    x        = "Target coverage (1 \u2212 \u03b1)",
    y        = "Proportion",
    color    = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Reading the plot: at 90% target coverage the model gives a confident singleton to roughly two-thirds of students. Raising coverage to 95% pushes more students into the uncertain zone. The empirical coverage line tracks the target closely, confirming the guarantee holds.

---

# Part 8: Cross-Conformal Prediction

Split conformal uses a separate calibration set that is not available for training. If training data is limited, this cost may matter. Cross-conformal prediction (CV+) addresses this by using $k$-fold cross-validation: each fold serves as the calibration set for a model trained on the remaining folds. The scores from all folds are pooled.

The coverage guarantee remains valid. The cost is that $k$ models must be fitted.

`probably` implements this via `int_conformal_cv()`, which requires a `fit_resamples` result with predictions saved:

```{r cv_conformal}
fit_conformal_cv <- TRUE

if (fit_conformal_cv) {
  cv_folds <- vfold_cv(ou_train, v = 10)

  # For each fold: fit the finalized workflow on the training rows,
  # predict on the held-out rows, compute nonconformity scores.
  # map_dfr pools all fold scores into a single tibble.
  cv_scores <- map_dfr(cv_folds$splits, function(split) {
    fold_train <- training(split)
    fold_cal   <- testing(split)

    fold_fit <- fit(final_xgb, fold_train)

    augment(fold_fit, fold_cal) %>%
      mutate(
        true_class_prob = if_else(result == "passed", .pred_passed, .pred_not_passed),
        score           = 1 - true_class_prob
      ) %>%
      select(result, score)
  })

  # Pooled quantile threshold across all fold calibration scores
  n_cv       <- nrow(cv_scores)
  q_level_cv <- min(ceiling((n_cv + 1) * (1 - alpha)) / n_cv, 1)
  q_hat_cv   <- quantile(cv_scores$score, probs = q_level_cv)

  cat("CV calibration scores:", n_cv, "\n")
  cat("CV score threshold:   ", round(q_hat_cv, 4), "\n")

  # Apply to test data using the model fit on all of ou_train (xgb_fit)
  cv_conformal_sets <- test_preds %>%
    mutate(
      include_passed     = (1 - .pred_passed)     <= q_hat_cv,
      include_not_passed = (1 - .pred_not_passed) <= q_hat_cv,
      covered = case_when(
        result == "passed"     & include_passed     ~ TRUE,
        result == "not_passed" & include_not_passed ~ TRUE,
        TRUE                                        ~ FALSE
      )
    )

  cv_conformal_sets %>%
    summarise(
      empirical_coverage = mean(covered),
      target_coverage    = 1 - alpha,
      n_test             = n()
    ) %>%
    print()
}
```


```{r}
cv_conformal_sets%>%
  select(result,.pred_class,.pred_passed,.pred_not_passed,include_passed,include_not_passed,covered)%>%
  group_by(result,covered)%>%
  count()

# Add derived columns used by both table and plot
cv_summary <- cv_conformal_sets %>%
  mutate(
    true_class_prob = if_else(result == "passed", .pred_passed, .pred_not_passed),
    set_size = include_passed + include_not_passed,
    set_type = case_when(
      set_size == 2 ~ "Uncertain (both)",
      set_size == 0 ~ "Anomaly (empty)",
      TRUE          ~ "Confident (singleton)"
    )
  )
```


```{r}
cv_summary %>%
  group_by(result, set_type) %>%
  summarise(
    n              = n(),
    pct            = n / nrow(cv_summary),
    mean_true_prob = mean(true_class_prob),
    pct_covered    = mean(covered),
    .groups        = "drop"
  ) %>%
  arrange(result, set_type) %>%
  mutate(across(c(pct, mean_true_prob, pct_covered), ~ round(.x, 3)))
```

```{r}
ggplot(cv_summary, aes(x = true_class_prob, fill = set_type)) +
  geom_histogram(bins = 40, color = "white", alpha = 0.85) +
  facet_wrap(~ result, scales = "free_y", labeller = label_both) +
  scale_fill_manual(
    values = c(
      "Confident (singleton)" = "steelblue",
      "Uncertain (both)"      = "firebrick",
      "Anomaly (empty)"       = "grey50"
    )
  ) +
  geom_vline(xintercept = 1 - q_hat_cv, linetype = "dashed",
             color = "grey30", linewidth = 0.8) +
  labs(
    title    = "Distribution of true-class predicted probability by set type",
    subtitle = paste0("CV+ threshold q\u0302 = ", round(q_hat_cv, 3),
                      "  |  ", (1 - alpha) * 100, "% target coverage"),
    x        = "Predicted probability of true class",
    y        = "Count",
    fill     = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```



The key difference is practical rather than statistical: CV+ uses all of the training data for fitting (pooled across folds) and is preferable when $n$ is small. With large datasets like OULAD, split conformal is usually sufficient and much faster.

---

# Part 9: Practical Takeaways

## When Is Conformal Prediction Most Valuable?

The most important insight is that conformal prediction converts a poorly-calibrated probability into an actionable, auditable uncertainty statement. Its value is highest when:

- **High stakes decisions depend on individual predictions.** An aggregate AUC tells you about the model overall; a prediction set tells you about this student.
- **The model may be miscalibrated.** XGBoost probabilities are often not well-calibrated without post-hoc adjustment. Conformal sets are calibrated by construction.
- **You need a formal guarantee to a stakeholder.** "At least 90% of these prediction sets contain the true outcome" is a statement you can defend. "The model's AUC is 0.85" is harder to translate to individual reliability.

## Limitations to Keep in Mind

**Exchangeability.** The guarantee requires calibration and test data to come from the same distribution. If the student population shifts — a new semester, a different delivery platform — the guarantee may not hold. This is the same as standard distribution shift, but conformal prediction makes the assumption explicit.

**Efficiency, not accuracy.** A model that assigns 0.5 probability to everyone will still have valid 90% coverage — the prediction sets will just always include both classes. Conformal prediction tells you the model is being honest about its uncertainty, but it doesn't fix a bad model.

**The calibration set size matters.** With $n = 200$ calibration observations, the finite-sample correction $\frac{n+1}{n}$ is 0.5%. With $n = 50$, the quantile estimate becomes noisy. Larger calibration sets give tighter, more reliable thresholds.

---

# Summary

| Property | Conformal Prediction | Standard Model Intervals |
|---|---|---|
| Coverage guarantee | Exact (finite-sample) | Asymptotic / model-dependent |
| Distributional assumptions | Exchangeability only | Usually parametric |
| Works for any model | Yes | No |
| Output for classification | Prediction *sets* | Single probability |
| Calibration data needed | Yes | No |
| Computationally cheap | Yes (split) / Moderate (CV+) | Yes |

The procedure in three lines: compute a nonconformity score for each calibration observation, find the empirical quantile at the target level, include any class whose score on the test observation falls at or below that threshold. What comes out is a prediction set for each student with a guaranteed coverage rate — an honest, auditable, individual-level uncertainty statement that requires no parametric assumptions about the model or the data.
