---
title: "Cross Conformal Prediction"
author: "Will Doyle"
date: "2025-01-09"
output: github_document
---

See: https://www.tidymodels.org/learn/models/conformal-regression/

Rina Foygel Barber, Emmanuel J. Candès, Aaditya Ramdas, Ryan J. Tibshirani "Predictive inference with the jackknife+," The Annals of Statistics, 49(1), 486-507, 2021

https://www.stat.berkeley.edu/~ryantibs/statlearn-s23/lectures/conformal.pdf


$$
\text{Minimize: } \underbrace{\frac{1}{2n} \sum_{i=1}^n (y_i - \hat{y}_i)^2}_{\text{OLS loss}} + \lambda \left( \underbrace{\frac{1-\alpha}{2} \sum_{j=1}^p \beta_j^2}_{\text{L2 penalty}} + \underbrace{\alpha \sum_{j=1}^p |\beta_j|}_{\text{L1 penalty}} \right)
$$

-   $\lambda$ (penalty): Controls the **overall strength** of regularization. Larger $\lambda$ shrinks coefficients more aggressively.
-   $\alpha$ (mixture): Determines the **blend** between L1 and L2 penalties:
    -   $\alpha = 0$: Pure Ridge (L2 penalty only)
    -   $\alpha = 1$: Pure Lasso (L1 penalty only)
    -   $0 < \alpha < 1$: Elastic Net (mix of both penalties).

```{r}
library(tidyverse)
library(tidymodels)
library(janitor)
library(probably)
```

## Load dataset

```{r}
hs<-read_csv("hsls_extract.csv")%>%clean_names()
```

## Data Cleaning

```{r}
hs <- hs %>%
  mutate(across(-x1txmtscor, ~ ifelse(. < 0, NA, .)))%>%
  drop_na()
```

```{r}
hs_split<-initial_split(hs)

hs_train<-training(hs_split)

hs_test<-testing(hs_split)
```

## Recipe

```{r}
hs_formula<-as.formula("x3tgpatot~.")

hs_rec<-recipe(hs_formula,data=hs_train)%>%
  update_role(x3tgpatot,new_role = "outcome")%>%
  step_other(all_nominal_predictors(),threshold = .01)%>%
  step_dummy(all_nominal_predictors())%>%
  step_zv(all_predictors())%>%
  step_normalize(all_predictors())
```

In tidymodels, elastic net is set using two paramters. Penalty sets $\lambda$ while mixture sets $\alpha$ in the above notation. Below, we'll just set these as two values that need to be tuned.

```{r}
hs_tune_model<- 
  linear_reg(penalty=tune(),mixture=tune())%>% 
  set_engine("glmnet")
```

We can use tidymodels defaults to set values for penalty and mixture. Note how penalty goes on a log scale, while mixture just goes through the steps from 0 to 1 in a sensible manner.

```{r}
enet_grid<-grid_regular(extract_parameter_set_dials(hs_tune_model) ,levels=10)
```

```{r}
hs_rs<-vfold_cv(hs_train,v = 20) 
```

Set the workflow, as usual.

```{r}
hs_wf<-workflow()%>%
  add_model(hs_tune_model)%>%
  add_recipe(hs_rec)
```

Then we can use `tune_grid` to run the model through the resampled data, using the grid supplied.

```{r}
hs_enet_tune_fit <- 
  hs_wf %>%
    tune_grid(hs_rs,grid=enet_grid)
```


## Selecting best parameters

Our workflow will be much the same as last time once we've identified the best parameters. The `select_best` will pull the values from the tuning process with best mean rmse.

```{r}
best_params<-select_best(hs_enet_tune_fit)
```

## Finalizing the workflow

We can then take those best parameters and plug them in as the values to be used in fitting to the full training data.

```{r}
final_enet_workflow <- hs_wf %>%
  finalize_workflow(best_params)
```


## Fitting to the full training data

And finally we can fit the model to get the values for the coefficients.


```{r}
ctrl <- control_resamples(save_pred = TRUE, extract = I)

final_rs_model<-final_enet_workflow%>%
  fit_resamples(hs_rs,control=ctrl)

```


```{r}
cv_int <- int_conformal_cv(final_rs_model)
```


```{r}
test_cv_res_enet <- 
  predict(cv_int, hs_test, level = 0.90) %>% 
  bind_cols(hs_test)
```

```{r}
coverage <- function(x) {
  x %>% 
    mutate(in_bound = .pred_lower <= y & .pred_upper >= y) %>% 
    summarise(coverage = mean(in_bound) * 100)
}
```

```{r}
test_cv_res_enet$y<-test_cv_res_enet$x3tgpatot
coverage(test_cv_res_enet)
```


```{r}
## Without conformal interval
  ggplot(test_cv_res_enet)+
  geom_point(aes(x=x3tgpatot,y=.pred),color="blue",alpha=.1)

## With Conformal Interval
  ggplot(test_cv_res_enet)+
  geom_point(aes(x=x3tgpatot,y=.pred),color="blue",alpha=.1,size=.5)+
  geom_point(aes(x=x3tgpatot,y=.pred_lower),color="red",alpha=.1,size=.5)+
   geom_point(aes(x=x3tgpatot,y=.pred_upper),color="red",alpha=.1,size=.5)
```


```{r}
final_model <- final_enet_workflow %>%
  fit(data = hs_train)
```

```{r}
last_fit_enet<-last_fit(final_enet_workflow,hs_split)

collect_metrics(last_fit_enet)

```


## XGBOOST

```{r}
xgb_spec <- boost_tree(
  trees = 100, 
  tree_depth = tune(), 
  min_n = tune(),
  loss_reduction = tune(), ## first three: model complexity
  sample_size = tune(), 
  mtry = tune(),         ## randomness
  learn_rate = tune() ## step size
) %>%
  set_engine("xgboost") %>%
  set_mode("regression")
```

```{r}
xgb_grid <- grid_space_filling(
  tree_depth(),
  min_n(),
  loss_reduction(), 
  sample_size = sample_prop(),
  finalize(mtry(), hs_train),
  learn_rate(),
  size = 30
)
```

```{r}
hs_xg_wf <- workflow() %>%
  add_recipe(hs_rec) %>%
  add_model(xgb_spec)
```

```{r}
fit_model<-FALSE

if(fit_model){
xg_hs_tune_res <- tune_grid(
  hs_xg_wf,
  grid=xgb_grid,
  resamples = hs_rs,
)
save(xg_hs_tune_res,file="xg_hs_tune_res.Rdata")
} else{
  load("xg_hs_tune_res.Rdata")
}
```

```{r}
best_param<- select_best(xg_hs_tune_res,metric =  "rmse")
```


```{r}
final_xg_workflow <- hs_xg_wf %>%
  finalize_workflow(best_param)
```

```{r}
final_rs_xg_model<-final_xg_workflow%>%
  fit_resamples(hs_rs,control=ctrl)
```


```{r}
cv_int <- int_conformal_cv(final_rs_model)
```

```{r}
test_cv_res_xg <- 
  predict(cv_int, hs_test, level = 0.90) %>% 
  bind_cols(hs_test)
```


```{r}
test_cv_res_xg$y<-test_cv_res_xg$x3tgpatot
coverage(test_cv_res_xg)
```


```{r}
## Without conformal interval
  ggplot(test_cv_res_enet)+
  geom_point(aes(x=x3tgpatot,y=.pred),color="blue",alpha=.1)

## With Conformal Interval
  ggplot(test_cv_res_enet)+
  geom_point(aes(x=x3tgpatot,y=.pred),color="blue",alpha=.1,size=.5)+
  geom_point(aes(x=x3tgpatot,y=.pred_lower),color="red",alpha=.1,size=.5)+
   geom_point(aes(x=x3tgpatot,y=.pred_upper),color="red",alpha=.1,size=.5)
```

```{r}
last_fit_xg<-last_fit(final_xg_workflow,hs_split)

collect_metrics(last_fit_xg)
```

