---
title: "Lecture 21: Downsampling for Rare Outcomes"
author: "Will Doyle"
date: "2026-04-16"
output: github_document
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(janitor)
library(themis)
library(patchwork)
```



# The Problem: When Accuracy Is Free

Most classifiers are implicitly optimized for accuracy. When one class is rare,
a model can achieve high accuracy by ignoring it entirely — predicting the
majority class for every observation. This is the null classifier (AKA Doyle Consulting), and it
sets a baseline that any useful model must beat.

Consider predicting whether a student finishes high school in the top quartile
of GPA. About 25% of students fall there by construction. A model that predicts
"not top quartile" for everyone is correct 75% of the time without learning
anything. Accuracy alone will not tell us whether our model is doing useful work.

**Downsampling**, **upsampling**, and **SMOTE** are three strategies for
rebalancing the training data so the model is forced to take the minority class
seriously. Each works differently and makes different assumptions.

The key constraint shared by all three: we only modify the **training** data.
The test set always reflects the true class distribution — otherwise our
evaluation metrics would be meaningless. The idea is to train the model one way,
but to test it on the held out full testing dataset.


# Data Setup

```{r load_data}
hs <- read_csv("hsls_extract.csv") %>%
  clean_names() %>%
  mutate(across(where(is.numeric), ~ replace(.x, .x %in% -9:-1, NA))) %>%
  drop_na()
```

## Create the Outcome

We'll define the top quartile of `x3tgpatot` as our positive class. Unlike the
top-decile example you may have seen before, the top quartile gives us a 25%
positive rate — rare enough to cause the null-classifier problem, but not so
extreme that the minority class is vanishingly small.

```{r create_outcome}
gpa_cutoff <- quantile(hs$x3tgpatot, 0.75)

hs <- hs %>%
  mutate(
    top_quartile = as_factor(if_else(x3tgpatot >= gpa_cutoff, "top_quartile", "other")),
    top_quartile = fct_relevel(top_quartile, "top_quartile")  # positive class first
  ) %>%
  select(-x3tgpatot)

hs %>%
  count(top_quartile) %>%
  mutate(proportion = n / sum(n))
```

The null classifier — predict "other" for everyone — would achieve ~75%
accuracy. Accuracy is therefore a misleading metric here. We'll use ROC AUC as
our primary tuning criterion, and report sensitivity and specificity alongside
it to understand the tradeoff the model is making.

---

# Train / Test Split

Stratify on the outcome to preserve the 75/25 class ratio in both splits.

```{r split}

hs_split <- initial_split(hs, prop = 0.75, strata = top_quartile)
hs_train <- training(hs_split)
hs_test  <- testing(hs_split)

hs_train %>%
  count(top_quartile) %>%
  mutate(proportion = n / sum(n))
```

---

# Four Recipes: Standard, Downsampled, Upsampled, and SMOTE

All four recipes share the same preprocessing steps. The only difference is
the resampling step added at the end of each.

**Downsampling** discards majority-class rows until both classes are equal.
No new information is created — you're just throwing away "other" students.

**Upsampling** duplicates minority-class rows until both classes are equal.
No majority-class data is lost, but the top-quartile students the model
sees are not new — they're exact copies of existing observations.

**SMOTE** (Synthetic Minority Oversampling TEchnique) creates *new* synthetic
minority-class observations by interpolating between existing ones. For each
top-quartile student, SMOTE finds its nearest neighbors in feature space among
other top-quartile students, then generates a new synthetic student at a random
point along the line connecting them. The result is a more varied minority-class
training set, at the cost of introducing observations that don't correspond to
any real student.

All three steps run **inside the resampling loop** during tuning — they only
touch the training fold of each resample. Validation folds and the final test
set always see the natural class distribution.

```{r recipes}
base_formula <- as.formula("top_quartile ~ .")

# Recipe WITHOUT any resampling
rec_standard <- recipe(base_formula, data = hs_train) %>%
  step_other(all_nominal_predictors(), threshold = 0.01) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_predictors())

# Downsample: reduce majority class to match minority
rec_downsample <- rec_standard %>%
  step_downsample(top_quartile, under_ratio = 1, seed = 2026)

# Upsample: replicate minority class to match majority
rec_upsample <- rec_standard %>%
  step_upsample(top_quartile, over_ratio = 1, seed = 2026)

# SMOTE: synthesize new minority-class observations by interpolation
rec_smote <- rec_standard %>%
  step_smote(top_quartile, over_ratio = 1, seed = 2026)
```

`under_ratio = 1` for downsampling means reduce the majority class until both
classes are equal. `over_ratio = 1` for upsampling and SMOTE means increase the
minority class until both classes are equal.

---

# Model Specification

Logistic elastic net with both `penalty` and `mixture` tuned. This is the same
specification used in earlier classification lectures.

```{r model_spec}
enet_spec <- logistic_reg(penalty = tune(), mixture = tune()) %>%
  set_engine("glmnet") %>%
  set_mode("classification")
```

# Workflows

```{r workflows}
wf_standard <- workflow() %>%
  add_recipe(rec_standard) %>%
  add_model(enet_spec)

wf_downsample <- workflow() %>%
  add_recipe(rec_downsample) %>%
  add_model(enet_spec)

wf_upsample <- workflow() %>%
  add_recipe(rec_upsample) %>%
  add_model(enet_spec)

wf_smote <- workflow() %>%
  add_recipe(rec_smote) %>%
  add_model(enet_spec)
```

---

# Resampling and Tuning

Monte Carlo CV with 25 resamples, stratified on the outcome. Latin hypercube
grid with 25 candidates — enough to map the penalty/mixture surface without
exhaustive search. All four workflows use the same folds and the same grid so
comparisons are fair.

```{r resamples}
set.seed(2026)

hs_folds <- mc_cv(hs_train, times = 25, prop = 0.75, strata = top_quartile)

enet_grid <- grid_space_filling(
  extract_parameter_set_dials(enet_spec),
  size = 25
)
```

```{r tune}
fit_model <- TRUE

if (fit_model) {
  tune_standard <- wf_standard %>%
    tune_grid(
      hs_folds,
      grid    = enet_grid,
      metrics = metric_set(roc_auc, sensitivity, specificity)
    )

  tune_downsample <- wf_downsample %>%
    tune_grid(
      hs_folds,
      grid    = enet_grid,
      metrics = metric_set(roc_auc, sensitivity, specificity)
    )

  tune_upsample <- wf_upsample %>%
    tune_grid(
      hs_folds,
      grid    = enet_grid,
      metrics = metric_set(roc_auc, sensitivity, specificity)
    )

  tune_smote <- wf_smote %>%
    tune_grid(
      hs_folds,
      grid    = enet_grid,
      metrics = metric_set(roc_auc, sensitivity, specificity)
    )

  save(tune_standard, tune_downsample, tune_upsample, tune_smote,
       file = here::here("scripts", "downsample_tune.Rdata"))
} else {
  load(here::here("scripts", "downsample_tune.Rdata"))
}
```

---

# Tuning Results

```{r compare_tuning}
bind_rows(
  tune_standard   %>% collect_metrics() %>% mutate(model = "Standard"),
  tune_downsample %>% collect_metrics() %>% mutate(model = "Downsampled"),
  tune_upsample   %>% collect_metrics() %>% mutate(model = "Upsampled"),
  tune_smote      %>% collect_metrics() %>% mutate(model = "SMOTE")
) %>%
  group_by(model, .metric) %>%
  slice_max(mean, n = 1) %>%
  select(model, .metric, mean, std_err) %>%
  arrange(.metric, model)
```

```{r plot_tuning}
bind_rows(
  tune_standard   %>% collect_metrics() %>% mutate(model = "Standard"),
  tune_downsample %>% collect_metrics() %>% mutate(model = "Downsampled"),
  tune_upsample   %>% collect_metrics() %>% mutate(model = "Upsampled"),
  tune_smote      %>% collect_metrics() %>% mutate(model = "SMOTE")
) %>%
  filter(.metric == "roc_auc") %>%
  ggplot(aes(x = penalty, y = mean, color = model)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE) +
  scale_x_log10() +
  scale_color_manual(values = c(
    "Standard"    = "steelblue",
    "Downsampled" = "firebrick",
    "Upsampled"   = "forestgreen",
    "SMOTE"       = "darkorchid"
  )) +
  labs(
    title    = "Cross-validated ROC AUC across penalty values",
    subtitle = "All four training strategies on the same folds and grid",
    x        = "Penalty (log scale)",
    y        = "ROC AUC",
    color    = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

# Final Fit and Test-Set Evaluation

```{r finalize}
best_standard   <- select_best(tune_standard,   metric = "roc_auc")
best_downsample <- select_best(tune_downsample, metric = "roc_auc")
best_upsample   <- select_best(tune_upsample,   metric = "roc_auc")
best_smote      <- select_best(tune_smote,      metric = "roc_auc")

final_standard <- wf_standard %>%
  finalize_workflow(best_standard) %>%
  last_fit(hs_split, metrics = metric_set(roc_auc, sensitivity, specificity, accuracy))

final_downsample <- wf_downsample %>%
  finalize_workflow(best_downsample) %>%
  last_fit(hs_split, metrics = metric_set(roc_auc, sensitivity, specificity, accuracy))

final_upsample <- wf_upsample %>%
  finalize_workflow(best_upsample) %>%
  last_fit(hs_split, metrics = metric_set(roc_auc, sensitivity, specificity, accuracy))

final_smote <- wf_smote %>%
  finalize_workflow(best_smote) %>%
  last_fit(hs_split, metrics = metric_set(roc_auc, sensitivity, specificity, accuracy))
```

```{r test_metrics}
bind_rows(
  collect_metrics(final_standard)   %>% mutate(model = "Standard"),
  collect_metrics(final_downsample) %>% mutate(model = "Downsampled"),
  collect_metrics(final_upsample)   %>% mutate(model = "Upsampled"),
  collect_metrics(final_smote)      %>% mutate(model = "SMOTE")
) %>%
  select(model, .metric, .estimate) %>%
  pivot_wider(names_from = model, values_from = .estimate) %>%
  rename(metric = .metric) %>%
  arrange(metric)
```

Reading across the rows:

- **ROC AUC** tends to be similar across all four strategies — resampling
  primarily shifts the operating point on the ROC curve, it doesn't reshape it.
- **Sensitivity** rises substantially for all three resampling strategies
  relative to the standard model. The model is now catching more true
  top-quartile students.
- **Specificity** falls correspondingly — the price of higher sensitivity is
  more false positives among the "other" students.
- **Accuracy** drops for the resampled models, because more observations are
  being sent across the 0.5 threshold into the positive class.

Downsampling, upsampling, and SMOTE tend to produce similar sensitivity/
specificity profiles. The differences between them are usually smaller than the
difference between any of them and the standard model.

---

# ROC Curves

```{r roc_curves}
bind_rows(
  final_standard   %>% collect_predictions() %>%
    roc_curve(truth = top_quartile, .pred_top_quartile) %>% mutate(model = "Standard"),
  final_downsample %>% collect_predictions() %>%
    roc_curve(truth = top_quartile, .pred_top_quartile) %>% mutate(model = "Downsampled"),
  final_upsample   %>% collect_predictions() %>%
    roc_curve(truth = top_quartile, .pred_top_quartile) %>% mutate(model = "Upsampled"),
  final_smote      %>% collect_predictions() %>%
    roc_curve(truth = top_quartile, .pred_top_quartile) %>% mutate(model = "SMOTE")
) %>%
  ggplot(aes(x = 1 - specificity, y = sensitivity, color = model)) +
  geom_line(linewidth = 1) +
  geom_abline(linetype = "dashed", color = "grey50") +
  scale_color_manual(values = c(
    "Standard"    = "steelblue",
    "Downsampled" = "firebrick",
    "Upsampled"   = "forestgreen",
    "SMOTE"       = "darkorchid"
  )) +
  labs(
    title    = "ROC curves on held-out test set",
    subtitle = "Standard vs. three resampling strategies  |  outcome: top quartile of HS GPA",
    x        = "1 \u2212 Specificity (False Positive Rate)",
    y        = "Sensitivity (True Positive Rate)",
    color    = NULL
  ) +
  coord_equal() +
  theme_minimal() +
  theme(legend.position = "bottom")
```

If the four curves overlap closely, it confirms that resampling strategy is
primarily a threshold-shifting intervention. The overall discrimination ability
of the model — encoded in the shape of the curve — depends much more on the
features and the model family than on which resampling strategy we used.

---

# Predicted Probability Distributions

Comparing the predicted probability distributions shows the mechanical effect
of each strategy most clearly.

```{r pred_prob_dist}
bind_rows(
  final_standard   %>% collect_predictions() %>% mutate(model = "Standard"),
  final_downsample %>% collect_predictions() %>% mutate(model = "Downsampled"),
  final_upsample   %>% collect_predictions() %>% mutate(model = "Upsampled"),
  final_smote      %>% collect_predictions() %>% mutate(model = "SMOTE")
) %>%
  ggplot(aes(x = .pred_top_quartile, fill = top_quartile)) +
  geom_histogram(bins = 40, alpha = 0.7, color = "white", position = "identity") +
  facet_wrap(~ model, ncol = 2) +
  scale_fill_manual(values = c("steelblue", "firebrick")) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "grey40") +
  labs(
    title    = "Predicted probability distributions on the test set",
    subtitle = "All three resampling strategies shift probabilities rightward relative to standard",
    x        = "Predicted P(top quartile)",
    y        = "Count",
    fill     = "True outcome"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The standard model compresses predicted probabilities toward zero — it has
learned that low predictions are usually safe. All three resampling strategies
spread the distribution further across the full range. The dashed line at 0.5
marks the default classification threshold. Notice how many more observations
cross it under the resampled models.

The differences *among* the three resampled distributions are usually subtle.
Upsampling and downsampling tend to produce nearly identical predicted
probability distributions because both achieve the same 50/50 class balance
in training — they just get there differently. SMOTE may produce a slightly
smoother distribution because synthetic interpolated observations fill gaps in
the minority class's feature space that exact duplicates would not.

---

# Key Takeaways

**All three strategies address the same problem differently.**
Downsampling discards majority-class rows. Upsampling duplicates minority-class
rows verbatim. SMOTE creates synthetic minority-class rows by interpolating
between existing ones. All three achieve the same goal — a more balanced
training signal — but make different tradeoffs around information loss and
data fabrication.

**Resampling is a threshold-shifting strategy, not a model-improvement strategy.**
The overall shape of the ROC curve (and therefore AUC) changes little. What
changes is where the model sits on that curve at the default 0.5 threshold:
higher sensitivity, lower specificity, lower raw accuracy.

**The test set is always the real-world distribution.** All resampling steps
run inside the resampling loop via the recipe. The test set used in `last_fit()`
is never touched. Evaluating on a resampled test set would give misleadingly
optimistic sensitivity estimates.

**SMOTE's advantage is most visible with small datasets.** Upsampling and
downsampling both reduce effective training set diversity — upsampling by
repeating the same minority observations, downsampling by discarding majority
ones. SMOTE introduces genuine variation in the minority class, which can
reduce overfitting to specific minority-class examples. With a large dataset
like HSLS, this advantage is usually small.

**SMOTE assumes linear interpolation is meaningful.** A synthetic student
halfway between two real students in feature space should be a plausible
student. This assumption holds reasonably well for continuous predictors but
can produce implausible combinations with highly categorical data. In education
datasets with many dummy-coded variables, inspect SMOTE results carefully.

**The right choice depends on dataset size and feature type.** With large data
and mixed features, downsampling is simple and hard to break. With small data
and mostly continuous features, SMOTE often edges ahead. Upsampling sits in
between — it retains all majority-class information but adds no new minority-class
information. When in doubt, try all three and let the validation metrics decide.
