---
title: "Quick intro to tidyverse/ggplot"
author: "Will Doyle"
date: "2025-01-09"
output: github_document
---

## Tidyverse/GGPlot


## Load libraries

```{r}
library(tidyverse)
library(tidymodels)
library(janitor)
```


## Load dataset
```{r}
hs<-read_csv("hsls_extract.csv")%>%
  clean_names()
```

```{r}
hs <- hs %>%
  mutate(across(-x1txmtscor, ~ ifelse(. < 0, NA, .)))%>%
  drop_na()
```

## Basics of Tidyverse

Tidyverse is an approach to programming in R. It's guided by four principles:

1. Reuse existing data structures.

2. Compose simple functions with the pipe.

3. Embrace functional programming.

4. Design for humans.

It even has a manifesto! Check it out: https://tidyverse.tidyverse.org/articles/manifesto.html

Here's an explanation of what's going in the code above, which uses a Tidyverse approach to replacing all negative values with missing, except for the variable `x1tmtscor`

`hs <- hs %>%`

The hs data frame is being updated in place.

The `%>%` operator (pipe) is used to pass the data frame through a series of transformations.

`mutate(across(-x1txmtscor, ~ ifelse(. < 0, NA, .)))`

`mutate()`: Used to create or modify columns in the data frame.

`across(-x1txmtscor, ~ ifelse(. < 0, NA, .))`

Applies a transformation to all columns except x1txmtscor:

`-x1txmtscor`: Excludes the column x1txmtscor from the operation.
`~ ifelse(. < 0, NA, .)` : A lambda function that replaces negative values (. < 0) in the selected columns with NA. The . refers to the current column values.

Result: Negative values in all columns (except x1txmtscor) are replaced with NA.

`drop_na()`

Removes all rows from the data frame that contain NA values in any column.

## Outcome variable

```{r}
hs%>%
  summarize(mean(x3tgpatot))
```

The `summarize` command "collapses" the data, using the operator supplied, in this case `mean`

## GGPlot

```{r}
hs%>%
ggplot(aes(x=x3tgpatot))+
  geom_hist()
```

GGPlot stands for grammer of graphics. The basic idea is that a graphic consistes of elements that are mapped onto a geometry via an aesthetic (eg x and y variables mapped onto a scatterplot). A description is here: https://ggplot2.tidyverse.org/


## Outcome variable as a function of independent variable

Below, I use a tidyverse approach and then a ggplot approach to show average GPA by levels of parental income. 


## Table
```{r}
hs%>%
  group_by(x1par1edu)%>%
  summarize(`Mean GPA`=mean(x3tgpatot))%>%
  arrange(`Mean GPA`)%>%
  rename(`Parental Education`=x1par1edu)
```


## Graphic
```{r}
hs%>%
  group_by(x1par1edu)%>%
  summarize(`Mean GPA`=mean(x3tgpatot))%>%
  mutate(`Parental Education`=fct_reorder(x1par1edu,`Mean GPA`))%>%
  ggplot(aes(x=`Parental Education`,y=`Mean GPA`,fill=`Parental Education`))+
  geom_col()+
  coord_flip()+
  theme_minimal()+
  theme(legend.position="bottom")
```


